### 2024-03-14 paper/paper.qmd
upload changes made to the qmd document.

---
title: "My title"
subtitle: "My subtitle if needed"
author: Ahnaf Alam
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(rstanarm)
library(purrr)
library(kableExtra)
data <- read_csv(here::here("data/analysis_data/data.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

data1 <- pivot_longer(data, cols = -date, names_to = "variable", values_to = "value")

data1|>
  ggplot(aes(x = date, y = value, color = variable)) +
  geom_line() 
```

```{r}
#| echo: false
#| warning: false
#| label: tbl-2
#| tbl-cap: "Cleaned data showing real expenditure by different categories"

columns_to_format <- names(data)[-1]

formatted_data <- data
formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) sprintf("%.2f", x))

formatted_data[1:12,] |> 
  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
                      "Healthcare", "Services"),
        booktabs = TRUE,
        linesep = "",
        align = "c",
        format.args = list(big.mark = ",")
        ) |>
  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
  kable_styling(font_size = 9, latex_options = "scale_down") |>
  footnote(general = "This only shows quarterly data from 2007 to 2009. The actual data set consists of quarterly data
           from 2007 to 2022",
           threeparttable = TRUE)
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

data |>
  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  theme_classic() + 
  labs(
    x = "Real increase in disposable income",
    y = "Real increase in food expenditure"
  )
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: false
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-03-14 paper/paper.qmd
changes made to tables

@@
   column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
-  kable_styling(font_size = 9, latex_options = "scale_down") |>
-  footnote(general = "This only shows quarterly data from 2007 to 2009. The actual data set consists of quarterly data
-           from 2007 to 2022",
-           threeparttable = TRUE)
+  kable_styling(font_size = 9, latex_options = "scale_down") 
 ```

### 2024-03-16 paper/paper.qmd
uploading first model

@@
 library(kableExtra)
+
 data <- read_csv(here::here("data/analysis_data/data.csv"))
@@
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-1
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
+
+data |>
+  ggplot(aes(x = date, y =income_expenditure)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "Date",
+       y = "Real disposable income",
+       )
+```
+
+```{r}
+#| label: fig-2
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
+
+data |>
+  ggplot(aes(x = date, y = food_expenditure)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "Date",
+       y = "Real food expenditure",
+       )
+```
+
+```{r}
+#| label: tbl-1
+#| tbl-cap: "Description of cleaned data"
+#| warning: false
+#| message: false
 #| echo: false
 
-data1 <- pivot_longer(data, cols = -date, names_to = "variable", values_to = "value")
+table1 <-
+  tibble(
+    "Variable"= c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
+                      "Healthcare", "Services"),
+    "Description" = c("Date of data collected",
+                      "Real expenditure in durable goods",
+                      "Real expenditure in non-durable goods",
+                      "Real food expenditure",
+                      "Real disposable income",
+                      "Real expenditure in healthcare",
+                      "Real expenditure in services")
+  )
 
-data1|>
-  ggplot(aes(x = date, y = value, color = variable)) +
-  geom_line() 
+table1 |>
+  kable(booktabs = TRUE) |>
+  column_spec(c(1, 2)) |>
+  kable_styling(font_size = 9, latex_options = "hold_position")
+
 ```
@@
 
+We are at the back end of covid-19 pandemic yet risks of another widespread epidemic is still rather underestimated.
+At the beginning of this pandemic, public health officials were rather caught off-guard with the emergence and virulence of this disease.
+As such, public were relayed lots of mixed messages regarding what they can do to mitigate the risk of infection and this confusion only added to public panic.
+Measures like Spain banning smoking in public were, or mandatory gloves Russia and Ukraine  were implemented, which may sound absurd right now, although at the time seemed reasonable.
+However, public health experts soon realized effectiveness of masks in slowing transmission rate.
+A meta-analysis, consisting of 6 studies from 4 different countries show that masks significantly reduce transmission risks (OR = 0.38, 95% CI: 0.21-0.69, I2 = 54.1%)  , and among health care workers, masks reduce transmission by as much as 70%.
+
 # Additional data details

### 2024-03-16 paper/paper.qmd
second model added

@@
 
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+first_model <- 
+  readRDS(file = here::here("models/model_1.rds"))
+
+modelsummary::modelsummary(
+  list(
+    "First Model" = first_model
+  ),
+  statistic = "mad",
+  fmt = 2
+)
+```
+
 The goal of our modelling strategy is twofold. Firstly,...
@@
 
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
 ```
@@
 
-pp_check(first_model) +
+set.seed(853)
+
+data|> 
+  slice_sample(n = 1000)
+
+pp_check(model_1) +
   theme_classic() +
@@
 
-posterior_vs_prior(first_model) +
+posterior_vs_prior(model_1) +
   theme_minimal() +
@@
 
-plot(first_model, "rhat")
 ```

### 2024-03-16 paper/paper.qmd
updated qmd file

@@
 library(kableExtra)
+library(ggridges)
 
@@
 ```{r}
+data |>
+  ggplot(aes(x = date)) +
+  geom_area(aes(y = healthcare_expenditure, fill = "Healthcare")) +
+  #geom_area(aes(y = nondurable_expenditure, fill = "Non-durable goods")) +
+  #geom_area(aes(y = durable_expenditure, fill = "Durable goods")) +
+  #geom_area(aes(y = services_expenditure, fill = "Services")) +
+  labs(
+    x = "Date",
+    y = "Expenditure by categories"
+  ) +
+  theme_minimal()
+```
+
+```{r}
+#| label: fig-4
+#| fig-cap: Relationship between wing length and w
+#| echo: false
+#| warning: false
+#| message: false
+
+data$date <- as.Date(data$date)
+
+data_long <- reshape2::melt(data, id.vars = "date")
+
+data_long |>
+  ggplot( aes(x = date, y = value, fill = variable)) +
+  geom_area() +
+  labs(title = "Contribution of Expenditure Categories Over Time",
+       x = "Date",
+       y = "Expenditure",
+       fill = "Expenditure Category") +
+  theme_minimal()
+
+data_long
+
+```
+
+```{r}
 #| echo: false
@@
 
+second_model <- 
+  readRDS(file = here::here("models/model_2.rds"))
+
 modelsummary::modelsummary(
   list(
-    "First Model" = first_model
+    "First Model" = first_model,
+    "Second Model" = second_model
   ),
@@
 Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
 
@@
   coord_flip()
+
+```
+
+```{r}
+#| eval: false
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckandposteriorvsprior2
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+
+pp_check(model_2) +
+  theme_classic() +
+  theme(legend.position = "bottom")
+
+posterior_vs_prior(model_2) +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  theme(legend.position = "bottom") +
+  coord_flip()
 ```
@@
 
+plot(second_model, "trace")
+
 ```

### 2024-03-19 paper/paper.qmd
updating models and graphs

--- 2024-03-16 paper/paper.qmd
@@
 library(kableExtra)
+library(wesanderson)
+library(viridis)
+library(gridExtra)
 
@@
 #| label: fig-1
-#| fig-cap: Relationship between wing length and width
+#| fig-cap: Levels of real disposable income, in billions of dollars, chained to 2017 prices
 #| echo: false
@@
 #| label: fig-2
-#| fig-cap: Relationship between wing length and width
+#| fig-cap: Levels of real consumption expenditure, in billions of dollars, chained to 2017 prices
 #| echo: false
@@
 ```{r}
+#| label: fig-3
+#| fig-cap: Expenditure for different categories, in billions of 2017 dollars
+#| echo: false
+#| warning: false
+#| message: false
+
+data$date <- as.Date(data$date)
+
+data_long <- reshape2::melt(data, id.vars = "date")
+
+newnames <- c("durable_expenditure" = "Durable goods",
+              "nondurable_expenditure" = "Nondurable goods",
+              "food_expenditure" = "Food",
+              "income_expenditure" = "Disposable income",
+              "healthcare_expenditure" = "Halthcare",
+              "services_expenditure" = "Services")
+
+col <- viridis(6)
+
+figure <- data_long |>
+  ggplot( aes(x = date, y = value, fill = variable)) +
+  geom_area() +
+  labs(title = "Contribution of Expenditure Categories Over Time",
+       x = "Date",
+       y = "Expenditure",
+       fill = "Categories") +
+  theme_minimal() +
+  scale_fill_manual(values = col, labels = newnames, name = "Categories")
+
+print(figure)
+
+```
+
+```{r}
 #| echo: false
@@
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-4
+#| fig-cap: Relationship between increases in disposable income and increases in food expenditure, between 2007 and 2022.
 #| echo: false
@@
 # Model
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-3
+#| tbl-cap: Summary results for both models
+
+first_model <- 
+  readRDS(file = here::here("models/model_1.rds"))
+
+second_model <- 
+  readRDS(file = here::here("models/model_2.rds"))
+
+modelsummary::modelsummary(
+  list(
+    "First Model" = first_model,
+    "Second Model" = second_model
+  ),
+  statistic = "mad",
+  fmt = 2
+)
+```
 
@@
 Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
 
@@
 ```{r}
-#| eval: false
 #| echo: false
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
+#| label: fig-5
 #| layout-ncol: 2
 #| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+#| fig-subcap: ["Posterior prediction check of first model", "Posterior prediction check of second model"]
 
@@
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
+pp_check(second_model) +
+  theme_classic() +
+  theme(legend.position = "bottom")
+
 ```
@@
 #| warning: false
-#| label: fig-stanareyouokay
+#| label: fig-6
 #| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
+#| fig-subcap: ["Trace plot of Model 1", "Trace plot of model 2"]
 #| layout-ncol: 2
@@
 
-plot(first_model, "rhat")
+plot(second_model, "trace")
+
 ```

### 2024-03-19 scripts/04-model.R

@@ -13,25 +13,46 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read_csv(here::here("data/analysis_data/data.csv"))
 
 ### Model data ####
-first_model <-
+model_1 <-
   stan_glm(
-    formula = flying_time ~ length + width,
+    formula = food_expenditure ~ income_expenditure,
     data = analysis_data,
     family = gaussian(),
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+    seed = 485
   )
+
+
+model_2 <-
+  stan_glm(
+    formula = food_expenditure ~ income_expenditure + durable_expenditure + nondurable_expenditure +
+      healthcare_expenditure + services_expenditure,
+    data = analysis_data,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 485
+  )
+
 
 
 #### Save model ####
 saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+  model_1,
+  file = "models/model_1.rds"
 )
+
+saveRDS(
+  model_2,
+  file = "models/model_2.rds"
+)
+
+
 
 

### 2024-03-23 paper/paper.qmd
data section completed

@@
 bibliography: references.bib
+toc: true
 ---
@@
 library(tidyverse)
-library(palmerpenguins)
 library(ggplot2)
 library(rstanarm)
-library(purrr)
 library(kableExtra)
-library(wesanderson)
 library(viridis)
-library(gridExtra)
+library(testthat)
+library(fredr)
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+## Software and R-packages
+
+We create this project using statistical software, `R` [@citeR].
+For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2].
+The data used in this paper comes from [@fredr] package.
+We further used `rstanarm` [@rstanarm] for modelling.
+Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes.
+
+## Incorporating FRED data
+
+The data comes from FRED or Federal Reserve Economic Data, which is an online database, consisting of hundreds of thousands time series economic data on both US national level and international level.
+From the database, we incorporated six different datasets using `fredr` package.
+These were titled:
+
+- Real personal consumption expenditure: Food
+- Real disposable personal income
+- Real personal consumption expenditure: Durable Goods
+- Real personal consumption expenditure: Nondurable Goods
+- Real personal consumption expenditure: Services
+- Real personal consumption expenditure: Healthcare
+
+A detailed description of what each of these datasets reports on can be found on @tbl-1.
+There are few key features that are present in all of the datasets.
+Firstly, we only incorporate data between 2007 and 2022 on a quarterly basis.
+We used as an 2007 as an anchor because data on durable/nondurable goods is only available from year and we wanted to model our data on 15 year period, hence 2022.
+
+All the datasets are also seasonally adjusted and chained to 2017 dollars.
+Seasonally adjusted time series eliminates effect of seasonal influences.
+Seasonal influences like strikes, abnormal weather patterns, or events Boxing day sale, can distort the real underlying movements in the business cycle and adjusting for these variation, provides us with much clearer understanding of the dataset from period to period.
+The datasets used 2017 price level as reference point.
+This adjusts for inflation across time, allowing us to accurately compare economic data over multiple periods.
+We further adjust for inflation by using real economic data, as opposed to nominal data.
+This enables us to create valid comparison groups, allowing us to compare a category with another category across time.
+
+## Dataset characteristics
+
+Each dataset contains five different columns, with key ones being date and value.
+The data variable reports on the specific day on which data was collected.
+With quarterly data, we only see data on the first day from the months of January, April, August and October.
+Although quarterly, FRED does not include data for quarter months of August, or indeed December.
+This is more due to convenience as some data becomes available only after end of the quarter and updating the database on before that is not productive.
+Lastly, value columns reports on expenditure in billions of US dollars.
+For a better understanding of how each category measure up against one another, please see @fig-2.
+@tbl-2 reports in cleaned data that is being used for modelling and analysis.
+
+## Measurement
+
+FRED does not collect data itself but relies on public and private organizations to provide the database with data.
+Except for first and last observations of the month, FRED ignores missing value when it average, sum and end-of-period aggregation [@fred].
+In this context, missing values often arise during statutory holidays, when federal offices are closed.
+On those weeks, FRED only reports data on 6 days of the week, excluding the holiday and end-of-period calculation will be conducted based whatever the corresponding days are in that month, minus one.
 
@@
 #| label: fig-1
-#| fig-cap: Levels of real disposable income, in billions of dollars, chained to 2017 prices
+#| fig-cap: Levels of real consumption expenditure and income expenditure, in billions of dollars, chained to 2017 prices
+#| fig-subcap: ["Food expenditure", "Income"]
 #| echo: false
@@
 #| message: false
+#| layout-ncol: 2
+
+data |>
+  ggplot(aes(x = date, y = food_expenditure)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "Date",
+       y = "Real food expenditure",
+       )
 
@@
 ```{r}
-#| label: fig-2
-#| fig-cap: Levels of real consumption expenditure, in billions of dollars, chained to 2017 prices
-#| echo: false
-#| warning: false
-#| message: false
-
-data |>
-  ggplot(aes(x = date, y = food_expenditure)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "Date",
-       y = "Real food expenditure",
-       )
-```
-
-```{r}
 #| label: tbl-1
-#| tbl-cap: "Description of cleaned data"
+#| tbl-cap: "Description of the data variables"
 #| warning: false
@@
   tibble(
-    "Variable"= c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
+    "Expenditure variable"= c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
                       "Healthcare", "Services"),
     "Description" = c("Date of data collected",
-                      "Real expenditure in durable goods",
-                      "Real expenditure in non-durable goods",
-                      "Real food expenditure",
-                      "Real disposable income",
-                      "Real expenditure in healthcare",
-                      "Real expenditure in services")
+                      "Durable goods are goods that are more for future consumption than immediate consumption. These types
+                      of goods provides utlity over a length of period. Examples include machinery, tools, appliances
+                      among others",
+                      "Durable goods are anything that are generally consumed within a short period of time. Examples
+                      include food, clothing, cosmetics etc",
+                      "Expenditure in food by all Americans in a time period",
+                      "Refers to total income that is availble to individuals for consumption after deducing taxes",
+                      "The category reports on total expenditure on healthcare services, including medical treatments,
+                      medicine cost, physician services among other services",
+                      "This category encomapasses a variety of services, inluding education, transportation,
+                      utilities, hospitality and many others")
   )
 
-table1 |>
-  kable(booktabs = TRUE) |>
-  column_spec(c(1, 2)) |>
-  kable_styling(font_size = 9, latex_options = "hold_position")
+table1 %>%
+  kable("latex", booktabs = TRUE) %>%
+  column_spec(1, width = "2cm") %>%
+  column_spec(2, width = "14cm") %>%
+  kable_styling(latex_options = "hold_position", font_size = 9, position = "center")
 
@@
 ```{r}
-#| label: fig-3
+#| label: fig-2
 #| fig-cap: Expenditure for different categories, in billions of 2017 dollars
@@
               "income_expenditure" = "Disposable income",
-              "healthcare_expenditure" = "Halthcare",
+              "healthcare_expenditure" = "Healthcare",
               "services_expenditure" = "Services")
@@
   scale_fill_manual(values = col, labels = newnames, name = "Categories")
-
 print(figure)
@@
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
 ```{r}
-#| label: fig-4
+#| label: fig-3
 #| fig-cap: Relationship between increases in disposable income and increases in food expenditure, between 2007 and 2022.
@@
 
-Talk way more about it.
-
 # Model
@@
   list(
-    "First Model" = first_model,
-    "Second Model" = second_model
+    "Simple linear" = first_model,
+    "Multiple linear" = second_model
   ),
@@
 #| warning: false
-#| label: fig-5
+#| label: fig-4
 #| layout-ncol: 2
 #| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check of first model", "Posterior prediction check of second model"]
+#| fig-subcap: ["Posterior prediction check of simple linear regression", "Posterior prediction check of multiple linear regression"]
 
@@
 #| warning: false
-#| label: fig-6
+#| label: fig-5
 #| fig-cap: "Checking the convergence of the MCMC algorithm"

### 2024-03-24 paper/paper.qmd
model section completed

@@
 ```
-
-# Model
-
 ```{r}
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+\newpage
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+# Model
+
+In this section, we briefly discuss Bayesian models that are being used in this analysis. Background details and model diagnostics can be found under [Appendix -@sec-model-details].
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Using `rstanarm` library, we evaluated two Bayesian model, with one being simple linear regression, and another being multiple linear regression.
+The simple linear regression explores whether an increase in income leads to increase in expenditure in food.
+Multiple linear regression evaluates the same topic, however, controlling for various other predictors.
+
+### Simple linear regression
+
+Define $y_i$ as the expenditure in food in year $i$. Then $income_i$, is level of disposable income in year $i$, both in billions of US dollars.
 
@@
 y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
+\mu_i &= \beta_0 + \beta_1 \times \mbox{income}_i \\
+\beta_0 &\sim \mbox{Normal}(0, 279) \\
+\beta_1 &\sim \mbox{Normal}(0, 0.17) \\
+\sigma &\sim \mbox{Exponential}(0.009)
 \end{align}
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+### Multiple linear regression
 
-### Model justification
+Define $y_i$ as the expenditure in food in year $i$.
+Then $income_i$, is level of disposable income in year $i$.
+Model further controls for durable goods with $durable_i$, non-durable goods with $nondurable_i$, levels of health care expenditure with $healthcare_i$ and levels of expenditure in services with $services_i$.
+All the variables are in billions of US dollars.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+\begin{align}
+y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 + \beta_1 \mbox{income}_i + \beta_2 \mbox{durable}_i + \beta_3 \mbox{nondurable}_i +  \beta_4 \mbox{healthcare}_i + \beta_5 \mbox{services}_i\\
+\beta_0 &\sim \mbox{Normal}(0, 279) \\
+\beta_1 &\sim \mbox{Normal}(0, 0.17) \\
+\beta_2 &\sim \mbox{Normal}(0, 0.84) \\
+\beta_3 &\sim \mbox{Normal}(0, 0.95) \\
+\beta_4 &\sim \mbox{Normal}(0, 1.25) \\
+\beta_5 &\sim \mbox{Normal}(0, 0.27) \\
+\sigma &\sim \mbox{Exponential}(0.009)
+\end{align}
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+## Model justification
+
+We expect a positive relationship between between income and food expenditure.
+@fig-3 shows that for an increase an disposable income, food expenditure increases by approximately equal amount.
+Further, @fig-1 compares both of these variables over time and we see similar patterns of growth, with steady exponential increase in between 2010 and 2020 and both measures veer off after 2020, presumably due to Covid-19 downturn.
+A paper by @parker1997household looks at data from Mexico and finds that income and expenditure are correlated.
+In fact, lower income uninsured groups reduces cash expenditure on health care during economic crisis.
+@mahadea2008economic further considers relationship between between happiness and incomes and finds that economic growth and increased income contributes to happiness.
+Therefore, based on exploratory data analysis and theories, we believe that there is positive relationship between food expenditure and income levels.
 
@@
 # Appendix {-}
-
-We are at the back end of covid-19 pandemic yet risks of another widespread epidemic is still rather underestimated.
-At the beginning of this pandemic, public health officials were rather caught off-guard with the emergence and virulence of this disease.
-As such, public were relayed lots of mixed messages regarding what they can do to mitigate the risk of infection and this confusion only added to public panic.
-Measures like Spain banning smoking in public were, or mandatory gloves Russia and Ukraine  were implemented, which may sound absurd right now, although at the time seemed reasonable.
-However, public health experts soon realized effectiveness of masks in slowing transmission rate.
-A meta-analysis, consisting of 6 studies from 4 different countries show that masks significantly reduce transmission risks (OR = 0.38, 95% CI: 0.21-0.69, I2 = 54.1%)  , and among health care workers, masks reduce transmission by as much as 70%.
 

### 2024-03-24 paper/paper.qmd
datasheet added

@@
 
+\newpage
+
 # Introduction
@@
 For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2].
-The data used in this paper comes from [@fredr] package.
+The data used in this paper comes from `fredr` [@fredr] package.
 We further used `rstanarm` [@rstanarm] for modelling.
@@
 @tbl-2 reports in cleaned data that is being used for modelling and analysis.
+Datasheet is available at @sec-appen.
 
@@
 
+## Datasheet {#sec-appen}
+
+**Motivation**
+
+*For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
+
+> It was created in order to analyze how food expenditure changes when income fluctuates.
+While all the datasets can be found individually on FRED website, they however, are not complied together and my datasets fills that void.
+The dataset complies different types of expenditure into one data frame, which I believe can help to answer the proposed question.
+
+*Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
+
+> Ahnaf Alam, an undergraduate student at University of Toronto.
+
+*Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
+
+> No direct funding was received for this project.
+
+*Any other comments?*
+
+> No.
+
+**Composition**
+
+*What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
+Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)?
+Please provide a description.*
+
+> Each row of the dataset composed on valuation in billions of US dollars, on a specific date. The data provides information of cumulative spending habits by Americans throughout the year.
+
+*How many instances are there in total (of each type, if appropriate)?*
+
+> There are about 366 instances in total.
+
+*Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
+If the dataset is a sample, then what is the larger set?
+Is the sample representative of the larger set (for example, geographic coverage)?
+If so, please describe how this representativeness was validated/verified.
+If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
+
+> The dataset is a sample, howeverm it isn't random.
+The larger dataset consiststs of all the observations for expenditure on a daily basis from the time datasets were made available to FRED database.
+In that sense, sample is representative of the larger dataset as we see the patterns of consumptions and expenditure over 25 year period match the patterns we see our sample.
+
+*What data does each instance consist of? “Raw” data (for example, unprocessed text or images) or features? In either case, please provide a description.*
+
+> Each instance consists of value of expenditure in billions of USD, across different categories.
+
+*Is there a label or target associated with each instance? If so, please provide a description*
+
+> Yes, the unique consists of specific date
+
+*Is any information missing from individual instances?
+If so, please provide a description, explaining why this information is missing (for example, because it was unavailable).
+This does not include intentionally removed information, but might include, for example, redacted text.*
+
+> There are no missing values in the dataset.
+
+*Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
+
+> Yes, using the 'year' column.
+
+*Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
+
+> No.
+
+*Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
+
+> No.
+
+*Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
+If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer?
+Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
+
+> The data is self-contained. The data will exist and won't change over time.
+
+*Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)?
+If so, please provide a description.*
+
+> No. These are publicly avaiable data, released by public organizations
+
+*Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
+
+> No.
+
+*Does the dataset identify any sub-populations (for example, by age, gender)?
+If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
+
+> No.
+
+*Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
+
+> No. The data reports on national level.
+
+*Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
+If so, please provide a description.*
+
+> No.
+
+*Any other comments?*
+
+> No.
+
+**Collection process**
+
+*How was the data associated with each instance acquired?
+Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)?
+If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified?
+If so, please describe how.*
+
+> The data was collected from FRED website, using `fredr` package. The data was not directly observable but based on audits, taxes and surveys.
+
+*What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
+How were these mechanisms or procedures validated?*
+
+> We downloaded the data using `fredr` package on R.
+
+*If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
+
+> The dataset was collected based on specific years on a quarterly basis.
+
+*Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
+
+> Ahnaf Alam and no one else.
+
+*Over what timeframe was the data collected?
+Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)?
+If not, please describe the timeframe in which the data associated with the instances was created.*
+
+> The dataframe was created over period of one week.
+
+*Were any ethical review processes conducted (for example, by an institutional review board)?
+If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
+
+> No.
+
+*Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
+
+> No. We relied on third-party website for in all cases.
+
+*Were the individuals in question notified about the data collection?
+If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
+
+> No.
+
+*Did the individuals in question consent to the collection and use of their data?
+If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
+
+> No.
+
+*If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?
+If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
+
+> Consent was not needed as we are dealing with data on country level.
+
+*Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted?
+If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
+
+> No.
+
+*Any other comments?*
+
+> No.
+
+**Preprocessing/cleaning/labelling**
+
+*Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)?
+If so, please provide a description.
+If not, you may skip the remaining questions in this section.*
+
+> Yes. The data was cleaned. There were no missing values in the dataset. From the raw data, we only selected columns that were pertinent to our question.
+
+*Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)?
+If so, please provide a link or other access point to the “raw” data.*
+
+> No. However, if one were to run the codes avaliable on `01-download_data.R` file, they can access the raw data.
+
+*Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
+
+> R was used.
+
+*Any other comments?*
+
+> No.
+
+**Uses**
+
+*Has the dataset been used for any tasks already? If so, please provide a description.*
+
+> Not that I am aware of.
+
+*Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
+
+> No.
+
+*What (other) tasks could the dataset be used for?*
+
+> we could use differnt types model to see how they compare to the ones we have performed in our paper.
+
+*Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
+For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)?
+If so, please provide a description.
+Is there anything a dataset consumer could do to mitigate these risks or harms?*
+
+> No.
+
+*Any other comments?*
+
+> No.
+
+**Distribution**
+
+*Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
+If so, please provide a description.*
+
+> The dataset will be avaiable on Github for later uses.
+
+*How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
+
+> It will be distributed through Github.
+
+*When will the dataset be distributed?*
+
+> The dataset is avaiable now.
+
+*Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?
+If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
+
+> No.
+
+*Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.8
+
+> None.
+
+*Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
+
+> No.
+
+*Any other comments?*
+
+> No.
+
+**Maintainence**
+
+*Who will be supporting/hosting/maintaining the dataset?*
+
+> Ahnaf Alam
+
+*How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
+
+> ahnaf.alam@mail.utoronto.ca
+
+*Is there an erratum? If so, please provide a link or other access point.*
+
+> No.
+
+*Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)?
+If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
+
+> No.
+
+*If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)?
+If so, please describe these limits and explain how they will be enforced.*
+
+> Not applicable.
+
+*Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
+
+> No.
+
+*If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
+If so, please provide a description.
+Will these contributions be validated/verified?
+If so, please describe how.
+If not, why not?
+Is there a process for communicating/distributing these contributions to dataset consumers?
+If so, please provide a description.*
+
+> Pull request on Github.
+
+*Any other comments?*
+
+> No.
+
 # Model details {#sec-model-details}

### 2024-03-29 paper/paper.qmd
added the data section

@@
 author: Ahnaf Alam
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure/tree/main"
 date: today
@@
   labs(x = "Date",
-       y = "Real food expenditure",
+       y = "Changes in real food expenditure",
        )
@@
   labs(x = "Date",
-       y = "Real disposable income",
+       y = "Changes in real disposable income",
        )
@@
   list(
-    "Simple linear" = first_model,
-    "Multiple linear" = second_model
+    "Model 1" = first_model,
+    "Model 2" = second_model
   ),
@@
 
-Our results are summarized in @tbl-modelresults.
+Our findings are outlined in @tbl-3.
+In simple linear regression, we observed an estimated intercept of 24.07 billion USD.
+Additionally, for every 1 billion USD increase in disposable income, there's a corresponding rise in food expenditure of $70 million USD.
+The small standard deviation suggests a low level of variability and uncertainty.
+
+In multiple linear regression, we found an estimated intercept of -97.40 billion USD.
+With each $1 billion USD increase in income, food expenditure increased by 20 million USD.
+Similar to the simple regression model, the standard deviation is negligibly small.
+
+Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit.
+Notably, the root mean square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.
+
+@fig-1 illustrates quarterly changes in disposable income and food expenditure from 2007 to 2022.
+Both variables show consistent growth trends, with 33.9% increase in disposable income and 32.5% increase in expenditure in food in this period.
+We see a notable surge in disposable income in 2020 due to COVID-era relief efforts.
+However, this increase was followed by a decline as restrictions eased, allowing for increased spending towards the end of 2021.
+
+@fig-2 displays the distribution of expenditure across categories, highlighting durable and non-durable goods as the largest expenditures.
+This aligns with expectations given the nature of these commodities.
+Following closely are food expenditure and disposable income, reflecting a strong correlation evident in @fig-3, where a clear linear relationship is observed with small standard deviations.
+Finally, healthcare and services expenditure are clustered together.
 

### 2024-03-29 scripts/04-model.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Models the 
+# Author: Ahnaf Alam
+# Date: March 29, 2024
+# Contact: ahnaf.alam@mail.utoronto.ca
+# Pre-requisites: Download and clean the data
+# Any other information needed?: None
 
 
 #### Workspace setup ####

### 2024-03-29 scripts/00-simulate_data.R
updating simulation file

@@ -1,19 +1,40 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Simulates the clean FRED dataset
+# Author: Ahnaf Alam
+# Date: March 29, 2024
+# Contact: ahnaf.alam@mail.utoronto.ca
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+
+set.seed(37373)
+
+dates <- seq(as.Date("2007-01-01"), by = "quarter", length.out = 60)
+
+# Without controls
+sim_data <- tibble(
+  Date = dates,
+  food = rnorm(60, mean = 950, sd = 150),
+  income = rnorm(60, mean = 1350, sd = 150)
+)
+
+# Adding in controls
+
+sim_data_controls <- tibble(
+  Date = dates,
+  food = rnorm(60, mean = 950, sd = 150),
+  income = rnorm(60, mean = 1350, sd = 150),
+  disposable = rnorm(60, mean = 2500, sd = 150),
+  nondisposable = rnorm(60, mean = 2350, sd = 150),
+  service = rnorm(60, mean = 2700, sd = 150),
+  health = rnorm(60, mean = 1900, sd = 150)
+)
 
 
 

### 2024-04-02 paper/paper.qmd
updating the paper

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "You have more money in your wallet. Does that mean you will spend extra on food?"
+subtitle: "I analyze the relationship between disposable income and food expenditure for American consumers"
 author: Ahnaf Alam
@@
 
+col <- viridis(6)
+
 newnames <- c("durable_expenditure" = "Durable goods",
@@
               "services_expenditure" = "Services")
-
-col <- viridis(6)
 
@@
 
-## First discussion point {#sec-first-point}
+@tbl-3 finds that for every billion dollar increase in disposable income, average expenditure in food goes up between 20 million USD and 70 million USD.
+This proves our hypothesis that there is a linear relationship between income and food consumption, with the models estimating economic significance of $70,000,000/1,000,000,000$ or 7% increase.
+In the upcoming section, we look to answer why we get a low economically significant value by assessing the role of savings
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+## Accounting for savings
 
-## Second discussion point
+A factor that may explain our low economic significance is that individuals choosing to save their income instead of spending.
+As income rises, consumption only rises by a fraction of the income [@krugman].
+It is plausible that as disposable increases, people rather chooses to save their additional income rather than to consume it.
+@mody2012precautionary analyzes the relationship between economic downturns and savings rate and finds that in advanced economies, income uncertainty is associated with higher household savings.
+@lusk2021consumer further looks to behavior of spending on food, specifically analyzing Covid-19 pandemic and finds that spending on food typically decreases during recessions.
+All this points suggests that increased level of saving, certainly could explain why people would choose to spend less on food and opt to save for future consumption.
 
-## Third discussion point
+## Interpreting consumer behavior
 
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+Current research suggests that higher income leads to higher expenditure on factors like durable goods.
+Consumers simply chooses to spend majority of their additional income on goods that provide utility over a considerable period of time than to spending on goods that will be consume over a single period.
+@lee1964income utilizes a linear probability function and finds that for durable goods, consumers typically adjust upwards in response to income increase and furthermore, they resist a downwards adjustment when income falls.
+This trend was visible during Covid-19 pandemic where increased disposable income led to higher than average spending on durable goods, including furniture and appliances, followed by motor vehicles and recreational goods [@tauber2021why].
+Therefore, it could simply be the case that food expenditure is less elastic to income shocks, compared to other factors like durable goods.
 

### 2024-04-06 paper/paper.qmd
finished the discussion section

@@
 
+## Factoring in elasticities {#sec-ela}
+
+Income and price elasticities of food is another factor that explains why low change in food expenditure for change in income.
+Income elasticity considers how sensitive are demand for food when income changes, while price elasticies looks at how change in price affects demand of food.
+@almaas2019income considers an experimental study in Kenya that randomized cash transfers to find income elasticity of food to 0.87.
+@purcell1967quantity expands on this and finds that income elasticies for items like poultry, fish and seafood, fruits were below 0.5.
+This establishes that food is normal commodity and as income rises, quantity demand for food also rises, albeit marginally.
+Further, @andreyeva2010impact conducted a meta-analysis on price elasticities of different food categories and found that essential times including fish, beef, poultry, fruits and vegetables ranged below 0.75.
+This shows that when price of food increases, demand for it goes down only slightly.
+This constitutes why we see only 7% economic significance as demand for food moderately increases when income goes and moderately decreases when prices of food goes up.
+Demand for food does not experience major shift due to changes in income.
+
+## Weaknesses and next steps
+
+The weaknesses of our model derives from the data we used.
+We used quarterly data from 2007 and 2022.
+We chose this range of dates because we were interested how the relationship between food expenditure and income changed over this time.
+This time period also captured the economic swings of Covid shocks.
+US and much of the world went into the pandemic in the fall of 2020.
+During the pandemic period, disposable income went up  sharply (see @fig-1) because social benefit programs put additional wealth into consumers pockets and moreover, consumers were not able to spend this wealth due to lockdown measures.
+Food expenditure, however, experienced a more diminished shock.
+This is because food is an essential commodity and one that you need for survival.
+Historical data on food elasticies also predict why expenditure on food does not change much as discussed on @sec-ela.
+This resulted in data where we see sharp and unpredictable swings in disposable income and predictable change in food expenditure and this distorted our data.
+
+This distortion is a problem this causes model to inaccurately estimate estimand.
+The true underlying relationship may not be as accurate as we would have hoped because the model yields biased estimates.
+It also diminishes external validity.
+Global pandemic is an unique situation and it introduced unique shocks that we do not foresee happening on a regular basis.
+This diminishes the generalizability and applicability of the model under normal circumstances.
+Lastly, policies around social benefit measures different from country to country during pandemic and further, lockdown measures differed between countries as well.
+It is plausible that economic swings US experienced may be different to that of other countries.
+Again, this reduces external validity.
+If we consider another country in the same 15 year period, our model may not be applicable as Covid-related policies were not the same as US's.
+
+In the next iteration of the paper, I would consider a time frame that excludes large scale shocks on global scale.
+As seen with Covid, it diminishes external validity and obscures estimand, which is what we are after.
+
 \newpage

### 2024-04-06 paper/datasheet_template.qmd
updated datasheet file

--- 2024-03-29 paper/paper.qmd
@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
-author: Ahnaf Alam
-thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure/tree/main"
+title: "Datasheet for data used in the paper, titled 'You have more money in your wallet. Does that mean you will spend extra on food?'"
+author: 
+  - Ahnaf Alam
+thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure "
 date: today
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
 format: pdf
@@
 bibliography: references.bib
-toc: true
 ---
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-library(tidyverse)
-library(ggplot2)
-library(rstanarm)
-library(kableExtra)
-library(viridis)
-library(testthat)
-library(fredr)
-
-data <- read_csv(here::here("data/analysis_data/data.csv"))
-
-```
-
-\newpage
-
-# Introduction
-
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
-
-The remainder of this paper is structured as follows. @sec-data....
-
-# Data {#sec-data}
-
-## Software and R-packages
-
-We create this project using statistical software, `R` [@citeR].
-For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2].
-The data used in this paper comes from `fredr` [@fredr] package.
-We further used `rstanarm` [@rstanarm] for modelling.
-Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes.
-
-## Incorporating FRED data
-
-The data comes from FRED or Federal Reserve Economic Data, which is an online database, consisting of hundreds of thousands time series economic data on both US national level and international level.
-From the database, we incorporated six different datasets using `fredr` package.
-These were titled:
-
-- Real personal consumption expenditure: Food
-- Real disposable personal income
-- Real personal consumption expenditure: Durable Goods
-- Real personal consumption expenditure: Nondurable Goods
-- Real personal consumption expenditure: Services
-- Real personal consumption expenditure: Healthcare
-
-A detailed description of what each of these datasets reports on can be found on @tbl-1.
-There are few key features that are present in all of the datasets.
-Firstly, we only incorporate data between 2007 and 2022 on a quarterly basis.
-We used as an 2007 as an anchor because data on durable/nondurable goods is only available from year and we wanted to model our data on 15 year period, hence 2022.
-
-All the datasets are also seasonally adjusted and chained to 2017 dollars.
-Seasonally adjusted time series eliminates effect of seasonal influences.
-Seasonal influences like strikes, abnormal weather patterns, or events Boxing day sale, can distort the real underlying movements in the business cycle and adjusting for these variation, provides us with much clearer understanding of the dataset from period to period.
-The datasets used 2017 price level as reference point.
-This adjusts for inflation across time, allowing us to accurately compare economic data over multiple periods.
-We further adjust for inflation by using real economic data, as opposed to nominal data.
-This enables us to create valid comparison groups, allowing us to compare a category with another category across time.
-
-## Dataset characteristics
-
-Each dataset contains five different columns, with key ones being date and value.
-The data variable reports on the specific day on which data was collected.
-With quarterly data, we only see data on the first day from the months of January, April, August and October.
-Although quarterly, FRED does not include data for quarter months of August, or indeed December.
-This is more due to convenience as some data becomes available only after end of the quarter and updating the database on before that is not productive.
-Lastly, value columns reports on expenditure in billions of US dollars.
-For a better understanding of how each category measure up against one another, please see @fig-2.
-@tbl-2 reports in cleaned data that is being used for modelling and analysis.
-Datasheet is available at @sec-appen.
-
-## Measurement
-
-FRED does not collect data itself but relies on public and private organizations to provide the database with data.
-Except for first and last observations of the month, FRED ignores missing value when it average, sum and end-of-period aggregation [@fred].
-In this context, missing values often arise during statutory holidays, when federal offices are closed.
-On those weeks, FRED only reports data on 6 days of the week, excluding the holiday and end-of-period calculation will be conducted based whatever the corresponding days are in that month, minus one.
-
-```{r}
-#| label: fig-1
-#| fig-cap: Levels of real consumption expenditure and income expenditure, in billions of dollars, chained to 2017 prices
-#| fig-subcap: ["Food expenditure", "Income"]
-#| echo: false
-#| warning: false
-#| message: false
-#| layout-ncol: 2
-
-data |>
-  ggplot(aes(x = date, y = food_expenditure)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "Date",
-       y = "Changes in real food expenditure",
-       )
-
-data |>
-  ggplot(aes(x = date, y =income_expenditure)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "Date",
-       y = "Changes in real disposable income",
-       )
-```
-
-```{r}
-#| label: tbl-1
-#| tbl-cap: "Description of the data variables"
-#| warning: false
-#| message: false
-#| echo: false
-
-table1 <-
-  tibble(
-    "Expenditure variable"= c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
-                      "Healthcare", "Services"),
-    "Description" = c("Date of data collected",
-                      "Durable goods are goods that are more for future consumption than immediate consumption. These types
-                      of goods provides utlity over a length of period. Examples include machinery, tools, appliances
-                      among others",
-                      "Durable goods are anything that are generally consumed within a short period of time. Examples
-                      include food, clothing, cosmetics etc",
-                      "Expenditure in food by all Americans in a time period",
-                      "Refers to total income that is availble to individuals for consumption after deducing taxes",
-                      "The category reports on total expenditure on healthcare services, including medical treatments,
-                      medicine cost, physician services among other services",
-                      "This category encomapasses a variety of services, inluding education, transportation,
-                      utilities, hospitality and many others")
-  )
-
-table1 %>%
-  kable("latex", booktabs = TRUE) %>%
-  column_spec(1, width = "2cm") %>%
-  column_spec(2, width = "14cm") %>%
-  kable_styling(latex_options = "hold_position", font_size = 9, position = "center")
-
-```
-
-```{r}
-#| label: fig-2
-#| fig-cap: Expenditure for different categories, in billions of 2017 dollars
-#| echo: false
-#| warning: false
-#| message: false
-
-data$date <- as.Date(data$date)
-
-data_long <- reshape2::melt(data, id.vars = "date")
-
-newnames <- c("durable_expenditure" = "Durable goods",
-              "nondurable_expenditure" = "Nondurable goods",
-              "food_expenditure" = "Food",
-              "income_expenditure" = "Disposable income",
-              "healthcare_expenditure" = "Healthcare",
-              "services_expenditure" = "Services")
-
-col <- viridis(6)
-
-figure <- data_long |>
-  ggplot( aes(x = date, y = value, fill = variable)) +
-  geom_area() +
-  labs(title = "Contribution of Expenditure Categories Over Time",
-       x = "Date",
-       y = "Expenditure",
-       fill = "Categories") +
-  theme_minimal() +
-  scale_fill_manual(values = col, labels = newnames, name = "Categories")
-print(figure)
-
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| label: tbl-2
-#| tbl-cap: "Cleaned data showing real expenditure by different categories"
-
-columns_to_format <- names(data)[-1]
-
-formatted_data <- data
-formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) sprintf("%.2f", x))
-
-formatted_data[1:12,] |> 
-  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
-                      "Healthcare", "Services"),
-        booktabs = TRUE,
-        linesep = "",
-        align = "c",
-        format.args = list(big.mark = ",")
-        ) |>
-  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
-  kable_styling(font_size = 9, latex_options = "scale_down") 
-```
-
-```{r}
-#| label: fig-3
-#| fig-cap: Relationship between increases in disposable income and increases in food expenditure, between 2007 and 2022.
-#| echo: false
-#| warning: false
-#| message: false
-
-data |>
-  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
-  geom_point(alpha = 0.5) +
-  geom_smooth(
-    method = "lm",
-    se = TRUE,
-    color = "black",
-    linetype = "dashed",
-    formula = y ~ x
-  ) +
-  theme_classic() + 
-  labs(
-    x = "Real increase in disposable income",
-    y = "Real increase in food expenditure"
-  )
-```
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: tbl-3
-#| tbl-cap: Summary results for both models
-
-first_model <- 
-  readRDS(file = here::here("models/model_1.rds"))
-
-second_model <- 
-  readRDS(file = here::here("models/model_2.rds"))
-
-modelsummary::modelsummary(
-  list(
-    "Model 1" = first_model,
-    "Model 2" = second_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
-
-\newpage
-
-# Model
-
-In this section, we briefly discuss Bayesian models that are being used in this analysis. Background details and model diagnostics can be found under [Appendix -@sec-model-details].
-
-## Model set-up
-
-Using `rstanarm` library, we evaluated two Bayesian model, with one being simple linear regression, and another being multiple linear regression.
-The simple linear regression explores whether an increase in income leads to increase in expenditure in food.
-Multiple linear regression evaluates the same topic, however, controlling for various other predictors.
-
-### Simple linear regression
-
-Define $y_i$ as the expenditure in food in year $i$. Then $income_i$, is level of disposable income in year $i$, both in billions of US dollars.
-
-\begin{align}
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \times \mbox{income}_i \\
-\beta_0 &\sim \mbox{Normal}(0, 279) \\
-\beta_1 &\sim \mbox{Normal}(0, 0.17) \\
-\sigma &\sim \mbox{Exponential}(0.009)
-\end{align}
-
-### Multiple linear regression
-
-Define $y_i$ as the expenditure in food in year $i$.
-Then $income_i$, is level of disposable income in year $i$.
-Model further controls for durable goods with $durable_i$, non-durable goods with $nondurable_i$, levels of health care expenditure with $healthcare_i$ and levels of expenditure in services with $services_i$.
-All the variables are in billions of US dollars.
-
-\begin{align}
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \mbox{income}_i + \beta_2 \mbox{durable}_i + \beta_3 \mbox{nondurable}_i +  \beta_4 \mbox{healthcare}_i + \beta_5 \mbox{services}_i\\
-\beta_0 &\sim \mbox{Normal}(0, 279) \\
-\beta_1 &\sim \mbox{Normal}(0, 0.17) \\
-\beta_2 &\sim \mbox{Normal}(0, 0.84) \\
-\beta_3 &\sim \mbox{Normal}(0, 0.95) \\
-\beta_4 &\sim \mbox{Normal}(0, 1.25) \\
-\beta_5 &\sim \mbox{Normal}(0, 0.27) \\
-\sigma &\sim \mbox{Exponential}(0.009)
-\end{align}
-
-## Model justification
-
-We expect a positive relationship between between income and food expenditure.
-@fig-3 shows that for an increase an disposable income, food expenditure increases by approximately equal amount.
-Further, @fig-1 compares both of these variables over time and we see similar patterns of growth, with steady exponential increase in between 2010 and 2020 and both measures veer off after 2020, presumably due to Covid-19 downturn.
-A paper by @parker1997household looks at data from Mexico and finds that income and expenditure are correlated.
-In fact, lower income uninsured groups reduces cash expenditure on health care during economic crisis.
-@mahadea2008economic further considers relationship between between happiness and incomes and finds that economic growth and increased income contributes to happiness.
-Therefore, based on exploratory data analysis and theories, we believe that there is positive relationship between food expenditure and income levels.
-
-# Results
-
-Our findings are outlined in @tbl-3.
-In simple linear regression, we observed an estimated intercept of 24.07 billion USD.
-Additionally, for every 1 billion USD increase in disposable income, there's a corresponding rise in food expenditure of $70 million USD.
-The small standard deviation suggests a low level of variability and uncertainty.
-
-In multiple linear regression, we found an estimated intercept of -97.40 billion USD.
-With each $1 billion USD increase in income, food expenditure increased by 20 million USD.
-Similar to the simple regression model, the standard deviation is negligibly small.
-
-Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit.
-Notably, the root mean square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.
-
-@fig-1 illustrates quarterly changes in disposable income and food expenditure from 2007 to 2022.
-Both variables show consistent growth trends, with 33.9% increase in disposable income and 32.5% increase in expenditure in food in this period.
-We see a notable surge in disposable income in 2020 due to COVID-era relief efforts.
-However, this increase was followed by a decline as restrictions eased, allowing for increased spending towards the end of 2021.
-
-@fig-2 displays the distribution of expenditure across categories, highlighting durable and non-durable goods as the largest expenditures.
-This aligns with expectations given the nature of these commodities.
-Following closely are food expenditure and disposable income, reflecting a strong correlation evident in @fig-3, where a clear linear relationship is observed with small standard deviations.
-Finally, healthcare and services expenditure are clustered together.
-
-# Discussion
-
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
-
-## Datasheet {#sec-appen}
+Extract of the questions from @gebru2021datasheets.
 
@@
 
-*For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
+1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
 
@@
 
-*Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
+2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
 
@@
 
-*Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
+3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
 
@@
 
-*Any other comments?*
+4. *Any other comments?*
 
@@
 
+1.
 *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
@@
 
-*How many instances are there in total (of each type, if appropriate)?*
+2. *How many instances are there in total (of each type, if appropriate)?*
 
@@
 
+3.
 *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
@@
 
-*What data does each instance consist of? “Raw” data (for example, unprocessed text or images) or features? In either case, please provide a description.*
+4. *What data does each instance consist of? “Raw” data (for example, unprocessed text or images) or features? In either case, please provide a description.*
 
@@
 
-*Is there a label or target associated with each instance? If so, please provide a description*
+5. *Is there a label or target associated with each instance? If so, please provide a description*
 
@@
 
+6.
 *Is any information missing from individual instances?
@@
 
-*Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
+7. *Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
 
@@
 
-*Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
+8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
 
@@
 
-*Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
+9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
 
@@
 
+10.
 *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
@@
 
+11.
 *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)?
@@
 
-*Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
+12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
 
@@
 
+13.
 *Does the dataset identify any sub-populations (for example, by age, gender)?
@@
 
-*Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
+14.
+*Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset?
+If so, please describe how.*
 
@@
 
+15.
 *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
@@
 
-*Any other comments?*
+16. *Any other comments?*
 
@@
 
+1.
 *How was the data associated with each instance acquired?
@@
 
+2.
 *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
@@
 
-*If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
+3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
 
@@
 
-*Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
+4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
 
@@
 
+5.
 *Over what timeframe was the data collected?
@@
 
+6.
 *Were any ethical review processes conducted (for example, by an institutional review board)?
@@
 
-*Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
+7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
 
@@
 
+8.
 *Were the individuals in question notified about the data collection?
@@
 
+9.
 *Did the individuals in question consent to the collection and use of their data?
@@
 
+10.
 *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?
@@
 
+11.
 *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted?
@@
 
-*Any other comments?*
+12. *Any other comments?*
 
@@
 
+1.
 *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)?
@@
 
+2.
 *Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)?
@@
 
-*Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
+3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
 
@@
 
-*Has the dataset been used for any tasks already? If so, please provide a description.*
+1. *Has the dataset been used for any tasks already? If so, please provide a description.*
 
@@
 
-*Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
+2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
 
@@
 
-*What (other) tasks could the dataset be used for?*
+3. *What (other) tasks could the dataset be used for?*
 
@@
 
+4.
 *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
@@
 
-*Any other comments?*
+5. *Any other comments?*
 
@@
 
+1.
 *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
@@
 
-*How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
+2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
 
@@
 
-*When will the dataset be distributed?*
+3. *When will the dataset be distributed?*
 
@@
 
+4.
 *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?
@@
 
+5.
 *Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
-If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.8
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
 
@@
 
+6.
 *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?
@@
 
-*Any other comments?*
+7. *Any other comments?*
 
@@
 
-*Who will be supporting/hosting/maintaining the dataset?*
+1. *Who will be supporting/hosting/maintaining the dataset?*
 
@@
 
-*How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
+2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
 
@@
 
-*Is there an erratum? If so, please provide a link or other access point.*
+3. *Is there an erratum? If so, please provide a link or other access point.*
 
@@
 
+4.
 *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)?
@@
 
+5.
 *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)?
@@
 
-*Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
+6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
 
@@
 
+7.
 *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
@@
 
-*Any other comments?*
+8. *Any other comments?*
 
 > No.
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-4
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check of simple linear regression", "Posterior prediction check of multiple linear regression"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-pp_check(second_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-5
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot of Model 1", "Trace plot of model 2"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(second_model, "trace")
-
-```
 

### 2024-04-06 paper/paper.qmd
updated datasheet file

@@
 author: Ahnaf Alam
-thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure/tree/main"
+thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure"
 date: today
@@
 *Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
-If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.8
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
 

### 2024-04-08 paper/paper.qmd
finished introduction and abstract

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+classoption: abstract
+abstract: "We estimated how food spending changes when disposable income increases.
+Using both a simple linear regression model and a multiple regression model, we found an economically insignificant increase of 20 million USD for every billion dollar increase in disposable income.
+At a policy level, our conclusion suggests that focusing solely on disposable income will not significantly affect food spending patterns.
+Policies aimed at stimulating consumer spending must consider other avenues."
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Wages have been going up in the US for the past thirty years, which usually means people have more money to spend after taxes.
+This extra money, called disposable income, is what's left over after taxes are taken out of a person's income.
+It's an important thing to look at because it shows how much money people have to spend.
+Research shows that when people have more disposable income, they tend to spend more money [@friend1946relationship].
+We're interested in how strong this connection is when it comes to one specific thing: food.
 
-The remainder of this paper is structured as follows. @sec-data....
+Food is something everyone needs to live.
+People might change what kinds of food they like over time, but everyone still needs to eat.
+In this project, we're looking at how closely linked disposable income is to how much people spend on food.
+This matters because if there's a strong connection, it suggests that when people make more money, they tend to spend more on basics like food.
+This could help predict how much people will spend on luxury items in the future.
+For instance, if we find a strong connection, it might mean that when people earn more, they spend a bigger chunk of their money on essentials like food, leaving less for luxury items.
+
+Numerous studies, notably summarized by Ernst Engel's law, suggest that as income and family size increase, spending on food goes up, but the proportion of income spent on food decreases [@lewbel2008engel].
+This means wealthier families tend to spend more on food, but it's a smaller part of their overall income.
+Taking inspiration from Engel's law, we want to see how food spending changes when we look at other spending categories instead of just income differences across families.
+
+Additionally, most past studies used income as the predictor variable, but we're using disposable income because we think it gives a better idea of overall wealth.
+The estimand of this paper is the measure of how much food spending changes (in billions of USD) when disposable income increases by a billion USD.
+
+In @sec-data, we discuss the data we used for this project, including information of the software and packages used.
+This is followed by a explanation of the models that we used, which can be found in @sec-model.
+In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income.
+We finish the project with weaknesses and potential areas of improvement in the future.
 
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-res}
 
@@
 
-# Discussion
+# Discussion {#sec-dis}
 

### 2024-04-12 paper/paper.qmd
updated the model selection section

@@
 toc: true
+margin-left: 25.4mm
+margin-right: 25.4mm
 ---
@@
 ```
-
-\newpage
 
@@
 
-The data comes from FRED or Federal Reserve Economic Data, which is an online database, consisting of hundreds of thousands time series economic data on both US national level and international level.
-From the database, we incorporated six different datasets using `fredr` package.
-These were titled:
+We gathered our data from FRED, the Federal Reserve Economic Data, an extensive online database housing a multitude of economic time series, both at the US national and international levels.
+Utilizing the `fredr` package, we integrated six distinct datasets into one dataframe.
+These datatsets were:
 
@@
 
-A detailed description of what each of these datasets reports on can be found on @tbl-1.
-There are few key features that are present in all of the datasets.
-Firstly, we only incorporate data between 2007 and 2022 on a quarterly basis.
-We used as an 2007 as an anchor because data on durable/nondurable goods is only available from year and we wanted to model our data on 15 year period, hence 2022.
+For a detailed breakdown of each dataset, refer to @tbl-1.
+Across all datasets, a few consistent features are notable.
+Firstly, we confined our data selection to the quarterly period spanning from 2007 to 2022.
+We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007.
+We were interested in a comprehensive 15-year modeling period, which culminated in 2022.
 
-All the datasets are also seasonally adjusted and chained to 2017 dollars.
-Seasonally adjusted time series eliminates effect of seasonal influences.
-Seasonal influences like strikes, abnormal weather patterns, or events Boxing day sale, can distort the real underlying movements in the business cycle and adjusting for these variation, provides us with much clearer understanding of the dataset from period to period.
-The datasets used 2017 price level as reference point.
-This adjusts for inflation across time, allowing us to accurately compare economic data over multiple periods.
-We further adjust for inflation by using real economic data, as opposed to nominal data.
-This enables us to create valid comparison groups, allowing us to compare a category with another category across time.
+All datasets underwent seasonal adjustment and were indexed to 2017 dollars.
+Seasonal adjustments mitigate the influence of periodic fluctuations, such as strikes or unusual weather events, which can obscure genuine economic trends.
+By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods.
+Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames.
 
@@
 
-Each dataset contains five different columns, with key ones being date and value.
-The data variable reports on the specific day on which data was collected.
-With quarterly data, we only see data on the first day from the months of January, April, August and October.
-Although quarterly, FRED does not include data for quarter months of August, or indeed December.
-This is more due to convenience as some data becomes available only after end of the quarter and updating the database on before that is not productive.
-Lastly, value columns reports on expenditure in billions of US dollars.
-For a better understanding of how each category measure up against one another, please see @fig-2.
-@tbl-2 reports in cleaned data that is being used for modelling and analysis.
-Datasheet is available at @sec-appen.
+Each dataset is structured with five distinct columns, the most significant being "date" and "value." The "date" column denotes the specific day of data collection.
+In the case of quarterly data, observations are restricted to the first day of January, April, July, and October.
+Notably, FRED does not provide data for August or December quarters.
+This omission is primarily due to practical considerations, as certain data may only become available after the quarter's end, rendering updates before then unproductive.
+
+Furthermore, the "value" column records expenditures in billions of US dollars.
+For a comparative analysis of each category, please refer to @fig-2.
+The cleaned data utilized for modeling and analysis is detailed in @tbl-2.
+The datasheet is accessible in @sec-appen for further reference.
 
@@
 
-FRED does not collect data itself but relies on public and private organizations to provide the database with data.
-Except for first and last observations of the month, FRED ignores missing value when it average, sum and end-of-period aggregation [@fred].
-In this context, missing values often arise during statutory holidays, when federal offices are closed.
-On those weeks, FRED only reports data on 6 days of the week, excluding the holiday and end-of-period calculation will be conducted based whatever the corresponding days are in that month, minus one.
+FRED, rather than directly collecting data, relies on contributions from both public and private organizations to populate its database.
+When aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].
+
+Missing values commonly occur during statutory holidays, when federal offices are closed.
+Consequently, during such weeks, FRED reports data for only six days, excluding the holiday.
+End-of-period calculations are adjusted accordingly, based on the remaining available days within the month, minus one.
 
@@
 
-We expect a positive relationship between between income and food expenditure.
-@fig-3 shows that for an increase an disposable income, food expenditure increases by approximately equal amount.
-Further, @fig-1 compares both of these variables over time and we see similar patterns of growth, with steady exponential increase in between 2010 and 2020 and both measures veer off after 2020, presumably due to Covid-19 downturn.
-A paper by @parker1997household looks at data from Mexico and finds that income and expenditure are correlated.
-In fact, lower income uninsured groups reduces cash expenditure on health care during economic crisis.
-@mahadea2008economic further considers relationship between between happiness and incomes and finds that economic growth and increased income contributes to happiness.
-Therefore, based on exploratory data analysis and theories, we believe that there is positive relationship between food expenditure and income levels.
+We anticipate a positive correlation between income and food expenditure.
+As depicted in @fig-3, an increase in disposable income corresponds to a similar rise in food expenditure.
+Additionally, @fig-1 compares these variables over time, revealing parallel growth patterns.
+Both exhibit a consistent, exponential increase between 2010 and 2020, followed by a divergence after 2020, likely due to the COVID-19 downturn.
+
+@sekhampu2012socio analyzes data from South Africa and employs a multiple linear regression model.
+The study concludes that household income, size, age, and employment status exert a significant positive influence on food expenditure.
+Likewise, @liu2019off utilizes Chinese data and a two-stage least squares estimator.
+Their findings reveal that for every 1,000 yuan increase in per-capita income, food expenditure rises by 61 yuan.
+Drawing from our exploratory data analysis and existing theories, we hypothesize a positive correlation between food expenditure and income levels.
+Therefore, we believe that both simple linear regression and multiple linear regression models are appropriate for addressing our research question.
 
@@
 
-@tbl-3 finds that for every billion dollar increase in disposable income, average expenditure in food goes up between 20 million USD and 70 million USD.
-This proves our hypothesis that there is a linear relationship between income and food consumption, with the models estimating economic significance of $70,000,000/1,000,000,000$ or 7% increase.
-In the upcoming section, we look to answer why we get a low economically significant value by assessing the role of savings
+@tbl-3 has discovered that for every billion-dollar rise in disposable income, the average expenditure on food increases by an amount ranging from 20 million USD to 70 million USD.
+This validates our hypothesis, suggesting a direct correlation between income and food consumption.
+According to the models, this translates to an economic significance of $70,000,000/$1,000,000,000, equivalent to a 7% increase.
+In the following section, we talk about the reasons behind relatively low economic significance, exploring the impact of savings, consumer behavior and elastiticies, followed by potential weaknesses in our paper.
 
@@
 
-A factor that may explain our low economic significance is that individuals choosing to save their income instead of spending.
-As income rises, consumption only rises by a fraction of the income [@krugman].
-It is plausible that as disposable increases, people rather chooses to save their additional income rather than to consume it.
-@mody2012precautionary analyzes the relationship between economic downturns and savings rate and finds that in advanced economies, income uncertainty is associated with higher household savings.
-@lusk2021consumer further looks to behavior of spending on food, specifically analyzing Covid-19 pandemic and finds that spending on food typically decreases during recessions.
-All this points suggests that increased level of saving, certainly could explain why people would choose to spend less on food and opt to save for future consumption.
+A factor that might clarify the low economic significance is that consumers tend to opt for saving their additional income rather than spending it.
+This aligns with existing economic theories, as suggested by [@krugman], which indicate that consumption increases only slightly in response to income growth, as individuals prefer to set aside funds for "rainy day".
+This tendency becomes apparent during periods of heightened economic uncertainty, when people tend to prioritize saving as a precautionary measure for the future.
+@mody2012precautionary explores economic downturns and savings rates further, finding that in advanced economies, income uncertainty correlates with higher household savings.
+Additionally, @lusk2021consumer explores spending behavior, particularly focusing on food expenditures during the COVID-19 pandemic, revealing a typical decrease in food spending during recessions.
+We see this concept playing out in @fig-2, where expenditure decreases on all consumption sectors, right around the start of COVID-19 pandemic in 2020.
+These findings suggest that an elevated level of saving could indeed be a reason why individuals opt to reduce their expenditure on food and prioritize saving for future consumption.
 
@@
 
-Current research suggests that higher income leads to higher expenditure on factors like durable goods.
-Consumers simply chooses to spend majority of their additional income on goods that provide utility over a considerable period of time than to spending on goods that will be consume over a single period.
-@lee1964income utilizes a linear probability function and finds that for durable goods, consumers typically adjust upwards in response to income increase and furthermore, they resist a downwards adjustment when income falls.
-This trend was visible during Covid-19 pandemic where increased disposable income led to higher than average spending on durable goods, including furniture and appliances, followed by motor vehicles and recreational goods [@tauber2021why].
-Therefore, it could simply be the case that food expenditure is less elastic to income shocks, compared to other factors like durable goods.
+Current research suggests that higher income tends to result in increased expenditure on durable goods, surpassing that of other consumable products.
+Consumers often allocate the majority of their additional income towards items that provide utility over an extended period rather than those consumed in a single instance.
+Using a linear probability function, @lee1964income showed that consumers typically increase their spending on durable goods in response to income growth and resist reducing it even when income decreases.
+This pattern was evident during the COVID-19 pandemic, where heightened disposable income led to above-average spending on durable goods such as furniture, appliances, motor vehicles, and recreational items [@tauber2021why].
+Therefore, it is plausible that food expenditure remains relatively stable compared to expenditures on durable goods, which may fluctuate more significantly in response to changes in income.
 
@@
 
-Income and price elasticities of food is another factor that explains why low change in food expenditure for change in income.
-Income elasticity considers how sensitive are demand for food when income changes, while price elasticies looks at how change in price affects demand of food.
-@almaas2019income considers an experimental study in Kenya that randomized cash transfers to find income elasticity of food to 0.87.
-@purcell1967quantity expands on this and finds that income elasticies for items like poultry, fish and seafood, fruits were below 0.5.
-This establishes that food is normal commodity and as income rises, quantity demand for food also rises, albeit marginally.
-Further, @andreyeva2010impact conducted a meta-analysis on price elasticities of different food categories and found that essential times including fish, beef, poultry, fruits and vegetables ranged below 0.75.
-This shows that when price of food increases, demand for it goes down only slightly.
-This constitutes why we see only 7% economic significance as demand for food moderately increases when income goes and moderately decreases when prices of food goes up.
-Demand for food does not experience major shift due to changes in income.
+The income and price elasticities of food are crucial factors in understanding why there's not much change in food spending despite changes in income.
+Income elasticity measures how much food demand shifts when income changes, while price elasticity assesses the impact of price changes on food demand.
+
+A study by @almaas2019income in Kenya, which randomly allocated cash transfers, found the income elasticity of food to be 0.87.
+Building on this, @purcell1967quantity found that income elasticities for items like poultry, fish, seafood, and fruits were below 0.5.
+This indicates that food is a normal commodity: as income increases, the demand for food also increases, but only slightly.
+
+Moreover, @andreyeva2010impact conducted a meta-analysis on the price elasticities of various food categories and discovered that essential items such as fish, beef, poultry, fruits, and vegetables had elasticities below 0.75.
+This suggests that when food prices rise, the decrease in demand is only slight.
+
+In essence, these findings show that there's only a modest increase in food demand as income rises and a moderate decrease as food prices go up.
+Consequently, changes in income don't lead to significant shifts in food demand.
 
@@
 
-The weaknesses of our model derives from the data we used.
-We used quarterly data from 2007 and 2022.
-We chose this range of dates because we were interested how the relationship between food expenditure and income changed over this time.
-This time period also captured the economic swings of Covid shocks.
-US and much of the world went into the pandemic in the fall of 2020.
-During the pandemic period, disposable income went up  sharply (see @fig-1) because social benefit programs put additional wealth into consumers pockets and moreover, consumers were not able to spend this wealth due to lockdown measures.
-Food expenditure, however, experienced a more diminished shock.
-This is because food is an essential commodity and one that you need for survival.
-Historical data on food elasticies also predict why expenditure on food does not change much as discussed on @sec-ela.
-This resulted in data where we see sharp and unpredictable swings in disposable income and predictable change in food expenditure and this distorted our data.
+Our model's weaknesses stem from the data we utilized—quarterly data spanning from 2007 to 2022.
+This period was chosen to explore changes in the relationship between food expenditure and income, however, it also captures economic impact of COVID-19, which was not our intention.
+With the pandemic's onset in late 2020, disposable income surged due to social benefit programs (see @fig-1-2), while food expenditure remained relatively stable due to its essential nature (see @fig-1-1).
+This led to erratic swings in income and predictable changes in food spending, distorting our data.
 
-This distortion is a problem this causes model to inaccurately estimate estimand.
-The true underlying relationship may not be as accurate as we would have hoped because the model yields biased estimates.
-It also diminishes external validity.
-Global pandemic is an unique situation and it introduced unique shocks that we do not foresee happening on a regular basis.
-This diminishes the generalizability and applicability of the model under normal circumstances.
-Lastly, policies around social benefit measures different from country to country during pandemic and further, lockdown measures differed between countries as well.
-It is plausible that economic swings US experienced may be different to that of other countries.
-Again, this reduces external validity.
-If we consider another country in the same 15 year period, our model may not be applicable as Covid-related policies were not the same as US's.
+This distortion poses challenges for accurately estimating our target variable or estimand.
+Unforeseen economic fluctuations during the pandemic may skew the true relationship we seek to understand, reducing the model's external validity.
+The unique nature of the pandemic's shocks further limits the model's generalizability, especially considering variations in social benefit policies and lockdown measures among different countries.
+Consequently, applying our model to another country within the same timeframe may yield unreliable results due to differing policy responses to COVID-19.
 
-In the next iteration of the paper, I would consider a time frame that excludes large scale shocks on global scale.
-As seen with Covid, it diminishes external validity and obscures estimand, which is what we are after.
+In future iterations, a narrower timeframe excluding such large-scale global shocks would be prudent.
+This approach aims to enhance the model's external validity and clarify the estimand we are pursuing.
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+We conducted posterior predictive checks for both models, with simple linear regression on @fig-4-1 and multiple linear regression on @fig-4-2.
+We wanted to understand the fit of the model and whether, can explain the observed data.
+For both graphs, $y$ is the observed data and $y_{\text{rep}}$ is the predicted data.
+@fig-4-1 shows that the Bayesian model fits the observed data well, however, there few spots where observed data does not track predicted data very well.
+We do not see this issue in multiple linear regression in @fig-4-2, where predicted and observed track each other well.
+Therefore, multiple linear regression may be more preferable than simple linear regression.
 
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+We use ran trace plots for both models and we are interested in understanding whether the algorithm ran into any issues [@Alexander_2023].
+Looking at the plots for both models, we do not encounter any out of the ordinary events.
 

### 2024-04-13 paper/paper.qmd
updates the test folder

@@
 
-Wages have been going up in the US for the past thirty years, which usually means people have more money to spend after taxes.
-This extra money, called disposable income, is what's left over after taxes are taken out of a person's income.
-It's an important thing to look at because it shows how much money people have to spend.
-Research shows that when people have more disposable income, they tend to spend more money [@friend1946relationship].
-We're interested in how strong this connection is when it comes to one specific thing: food.
+Over the past thirty years in the US, wages have consistently risen, while tax rates have remained relatively stable.
+This has resulted in people having more disposable income, which is the money left after taxes are deducted from one's income.
+Disposable income is crucial to consider as it reflects the actual amount of money available for spending.
+Studies indicate that when individuals have higher disposable income, they generally spend more [@friend1946relationship].
+We are particularly interested in examining this relationship in the context of food.
 
@@
 In this project, we're looking at how closely linked disposable income is to how much people spend on food.
-This matters because if there's a strong connection, it suggests that when people make more money, they tend to spend more on basics like food.
-This could help predict how much people will spend on luxury items in the future.
+This matters because if there's a strong connection, it suggests that when people make more money, they tend to spend more on basic goods and services like food.
+Further, this could help predict how much people will spend on luxury items in the future.
 For instance, if we find a strong connection, it might mean that when people earn more, they spend a bigger chunk of their money on essentials like food, leaving less for luxury items.
 
-Numerous studies, notably summarized by Ernst Engel's law, suggest that as income and family size increase, spending on food goes up, but the proportion of income spent on food decreases [@lewbel2008engel].
-This means wealthier families tend to spend more on food, but it's a smaller part of their overall income.
-Taking inspiration from Engel's law, we want to see how food spending changes when we look at other spending categories instead of just income differences across families.
-
+Numerous studies, notably summarized by Ernst Engel's law, suggest that as income and family size increase, spending on food goes up.
+However, the proportion of income spent on food decreases [@lewbel2008engel].
+This means wealthier families tend to spend more on food, but it is a smaller part of their overall income.
+Taking inspiration from Engel's law, we want to see how food spending changes when we account for other spending categories instead of just income differences across families.
 Additionally, most past studies used income as the predictor variable, but we're using disposable income because we think it gives a better idea of overall wealth.

### 2024-04-13 scripts/03-test_data.R

@@ -9,6 +9,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(testthat)
+library(lubridate)
 
 
 #### Test data ####
@@ -56,6 +57,35 @@ if (all(diff(data$date) > 0)) {
 } else {
   print("Dates are not in chronological order.")
 }
+
+
+#6.Test to see if all the quarterly data are in quarter months or not i.e. January, April, August, December
+
+q_dates <- function(dates) {
+  months <- month(dates)
+  q_months <-  c(1,4,8,12)
+  all(months %in% q_months)
+}
+
+test_that("Check quarterly dates", {
+  expect_true(q_dates(data$date))
+})
+
+#7. We check there are data for 4 distinct months for each year
+
+check_four_months <- function(data) {
+  data %>%
+    mutate(year = lubridate::year(date), 
+           month = lubridate::month(date)) |>
+    group_by(year) |>
+    summarise(num_months = n_distinct(month)) |>
+    mutate(check = num_months == 4)
+}
+
+
+
+
+
 
 
 

### 2024-04-14 paper/paper.qmd
updated the README

@@
 ---
-title: "You have more money in your wallet. Does that mean you will spend extra on food?"
-subtitle: "I analyze the relationship between disposable income and food expenditure for American consumers"
+title: "You have more money in your wallet. This does not mean you will spend extra on food"
+subtitle: "Examining the relationship between disposable income and food expenditure for American consumers"
 author: Ahnaf Alam
@@
 classoption: abstract
-abstract: "We estimated how food spending changes when disposable income increases.
-Using both a simple linear regression model and a multiple regression model, we found an economically insignificant increase of 20 million USD for every billion dollar increase in disposable income.
-At a policy level, our conclusion suggests that focusing solely on disposable income will not significantly affect food spending patterns.
-Policies aimed at stimulating consumer spending must consider other avenues."
+abstract: "We examined the relationship between changes in food spending and increases in disposable income.
+Utilizing quarterly country-level data from the Federal Reserve Economic Data (FRED) spanning a 15-year period from 2007 to 2022, we employed both a simple linear regression model and a multiple regression model.
+Our analysis revealed a economically insignificant increase of $20 million in food spending for every billion-dollar rise in disposable income.
+From a policy perspective, our findings indicate that solely targeting disposable income is unlikely to substantially impact food spending behaviors.
+Therefore, policies designed to boost consumer spending should explore alternative approaches."
 format: pdf
@@
 
-We gathered our data from FRED, the Federal Reserve Economic Data, an extensive online database housing a multitude of economic time series, both at the US national and international levels.
+We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels.
 Utilizing the `fredr` package, we integrated six distinct datasets into one dataframe.
@@
 
-FRED, rather than directly collecting data, relies on contributions from both public and private organizations to populate its database.
-When aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].
+FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S.
+Census Bureau, Bureau of Labor Statistics, and U.S.
+Bureau of Economic Analysis (BEA), among others.
+These datasets are aggregated at the country level.
+We use national-level data because it provides a broader perspective on general economic trends in the U.S.
+All five datasets we have used originate from the BEA.
+The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies.
+Upon receiving these datasets, BEA makes adjustments to the raw survey data.
+For instance, they fill in missing gaps in the datasets.
+For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from nonmerchant sellers.
+Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.
+
+FRED collaborates with both public and private organizations, such as the BEA, to populate its comprehensive database.
+The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S.
+economy, eliminating the need to search various agencies individually for data.
+However, FRED does make further adjustments to BEA datasets before releasing them.
+For example, when aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].
 
@@
 #| label: tbl-2
-#| tbl-cap: "Cleaned data showing real expenditure by different categories"
+#| tbl-cap: "Cleaned data showing real expenditure by different categories (in billions of 2017 US dollars)"
 
@@
   column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
-  kable_styling(font_size = 9, latex_options = "scale_down") 
+  kable_styling(font_size = 9, latex_options = "scale_down")  |>
+  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the of the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
 ```
@@
 y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \times \mbox{income}_i \\
+\mu_i &= \beta_0 + \beta_1 \times \mbox{income}_i + \epsilon\\
 \beta_0 &\sim \mbox{Normal}(0, 279) \\
 \beta_1 &\sim \mbox{Normal}(0, 0.17) \\
-\sigma &\sim \mbox{Exponential}(0.009)
+\sigma &\sim \mbox{Exponential}(0.009) \\
 \end{align}
@@
 y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \mbox{income}_i + \beta_2 \mbox{durable}_i + \beta_3 \mbox{nondurable}_i +  \beta_4 \mbox{healthcare}_i + \beta_5 \mbox{services}_i\\
+\mu_i &= \beta_0 + \beta_1 \mbox{income}_i + \beta_2 \mbox{durable}_i + \beta_3 \mbox{nondurable}_i +  \beta_4 \mbox{healthcare}_i + \beta_5 \mbox{services}_i + \epsilon\\
 \beta_0 &\sim \mbox{Normal}(0, 279) \\
@@
 Drawing from our exploratory data analysis and existing theories, we hypothesize a positive correlation between food expenditure and income levels.
-Therefore, we believe that both simple linear regression and multiple linear regression models are appropriate for addressing our research question.
+Therefore, we believe linear models would be appropriate for modelling our hypothesis.
+
+We utilize two linear models in our analysis: simple linear regression and multiple regression.
+Simple linear regression enables us to easily quantify the linear association between food expenditure and disposable income, as illustrated by @fig-3.
+However, this model has limitations.
+Most notably, it does not account for confounding variables, and it requires us to make unrealistic assumptions about the independence of the error term ($\epsilon$) and the predictor variable, income.
+This assumption implies that food expenditure and income are entirely independent of each other, which is problematic.
+Factors like economic downturns, income levels, and other variables can correlate with our model variables.
+
+To address these issues, we employ a multiple linear regression model.
+This model helps us control for potential confounding variables, such as expenditure on non-durable goods, which may influence consumer expenditure on food.
+Using multiple linear regression allows us to better isolate the effect of income on food expenditure.
+
+Between the two models, the multiple linear regression model may be more preferable than the simple linear regression model.
+Simple linear regression requires unrealistic assumptions that the multiple linear regression model relaxes.
+Furthermore, posterior predictive checks conducted for both models suggest that the multiple regression model may offer a better fit than the simple linear regression model.
+For more details, please refer to @sec-model-details.
 

### 2024-04-16 paper/paper.qmd
updated parquet

@@
 title: "You have more money in your wallet. This does not mean you will spend extra on food"
-subtitle: "Examining the relationship between disposable income and food expenditure for American consumers"
+subtitle: "Examining the relationship between disposable income and food expenditure for American consumers from 2007-2022"
 author: Ahnaf Alam
@@
 In @sec-data, we discuss the data we used for this project, including information of the software and packages used.
+This is followed any a explanation of the models that we used, which can be found in @sec-model.
+In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income.
+We finish the project with weaknesses and potential areas of improvement in the future.
+
+In @sec-data, we discuss the data we used for this project, including information of the software and packages used.
 This is followed by a explanation of the models that we used, which can be found in @sec-model.
@@
 We finish the project with weaknesses and potential areas of improvement in the future.
+Datasheet and model checks can be found under [Appendix -@sec-model-details].
 
 # Data {#sec-data}
-
-## Software and R-packages
-
-We create this project using statistical software, `R` [@citeR].
-For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2].
-The data used in this paper comes from `fredr` [@fredr] package.
-We further used `rstanarm` [@rstanarm] for modelling.
-Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes.
-
-## Incorporating FRED data
-
-We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels.
-Utilizing the `fredr` package, we integrated six distinct datasets into one dataframe.
-These datatsets were:
-
-- Real personal consumption expenditure: Food
-- Real disposable personal income
-- Real personal consumption expenditure: Durable Goods
-- Real personal consumption expenditure: Nondurable Goods
-- Real personal consumption expenditure: Services
-- Real personal consumption expenditure: Healthcare
-
-For a detailed breakdown of each dataset, refer to @tbl-1.
-Across all datasets, a few consistent features are notable.
-Firstly, we confined our data selection to the quarterly period spanning from 2007 to 2022.
-We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007.
-We were interested in a comprehensive 15-year modeling period, which culminated in 2022.
-
-All datasets underwent seasonal adjustment and were indexed to 2017 dollars.
-Seasonal adjustments mitigate the influence of periodic fluctuations, such as strikes or unusual weather events, which can obscure genuine economic trends.
-By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods.
-Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames.
-
-## Dataset characteristics
-
-Each dataset is structured with five distinct columns, the most significant being "date" and "value." The "date" column denotes the specific day of data collection.
-In the case of quarterly data, observations are restricted to the first day of January, April, July, and October.
-Notably, FRED does not provide data for August or December quarters.
-This omission is primarily due to practical considerations, as certain data may only become available after the quarter's end, rendering updates before then unproductive.
-
-Furthermore, the "value" column records expenditures in billions of US dollars.
-For a comparative analysis of each category, please refer to @fig-2.
-The cleaned data utilized for modeling and analysis is detailed in @tbl-2.
-The datasheet is accessible in @sec-appen for further reference.
-
-## Measurement
-
-FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S.
-Census Bureau, Bureau of Labor Statistics, and U.S.
-Bureau of Economic Analysis (BEA), among others.
-These datasets are aggregated at the country level.
-We use national-level data because it provides a broader perspective on general economic trends in the U.S.
-All five datasets we have used originate from the BEA.
-The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies.
-Upon receiving these datasets, BEA makes adjustments to the raw survey data.
-For instance, they fill in missing gaps in the datasets.
-For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from nonmerchant sellers.
-Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.
-
-FRED collaborates with both public and private organizations, such as the BEA, to populate its comprehensive database.
-The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S.
-economy, eliminating the need to search various agencies individually for data.
-However, FRED does make further adjustments to BEA datasets before releasing them.
-For example, when aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].
-
-Missing values commonly occur during statutory holidays, when federal offices are closed.
-Consequently, during such weeks, FRED reports data for only six days, excluding the holiday.
-End-of-period calculations are adjusted accordingly, based on the remaining available days within the month, minus one.
-
-```{r}
-#| label: fig-1
-#| fig-cap: Levels of real consumption expenditure and income expenditure, in billions of dollars, chained to 2017 prices
-#| fig-subcap: ["Food expenditure", "Income"]
-#| echo: false
-#| warning: false
-#| message: false
-#| layout-ncol: 2
-
-data |>
-  ggplot(aes(x = date, y = food_expenditure)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "Date",
-       y = "Changes in real food expenditure",
-       )
-
-data |>
-  ggplot(aes(x = date, y =income_expenditure)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "Date",
-       y = "Changes in real disposable income",
-       )
-```
 
@@
   kable_styling(latex_options = "hold_position", font_size = 9, position = "center")
+```
 
+## Software and R-packages
+
+We create this project using statistical software, `R` [@citeR].
+For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2].
+The data used in this paper comes from `fredr` [@fredr] package.
+We further used `rstanarm` [@rstanarm] and `modelsummary` [@ms] for modelling.
+Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes.
+For editing texts, we used `rspell` library [@rspell].
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: tbl-2
+#| tbl-cap: "Cleaned data showing real expenditure by different categories (in billions of 2017 US dollars)"
+
+columns_to_format <- names(data)[-1]
+
+formatted_data <- data
+formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) formatC(x, format = "f", digits = 2, big.mark = ","))
+
+formatted_data[1:12,] |> 
+  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
+                      "Healthcare", "Services"),
+        booktabs = TRUE,
+        linesep = "",
+        align = "c",
+        format = "latex") |> 
+  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
+  kable_styling(font_size = 9, latex_options = "scale_down")  |>
+  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
+```
+
+## Incorporating FRED data
+
+We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels.
+Utilizing the `fredr` package, we integrated six distinct datasets into one data frame.
+These datasets were:
+
+- Real personal consumption expenditure: Food
+- Real disposable personal income
+- Real personal consumption expenditure: Durable Goods
+- Real personal consumption expenditure: Nondurable Goods
+- Real personal consumption expenditure: Services
+- Real personal consumption expenditure: Healthcare
+
+For a detailed breakdown of each dataset, refer to @tbl-1.
+Across all datasets, a few consistent features are notable.
+Firstly, we confined our data selection to the quarterly period spanning from 2007 to 2022.
+We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007.
+We were interested in a comprehensive 15-year modeling period, which culminated in 2022.
+
+All datasets underwent seasonal adjustment and were indexed to 2017 dollars.
+Seasonal adjustments mitigate the influence of periodic fluctuations, such as strikes or unusual weather events, which can obscure genuine economic trends.
+By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods.
+Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames
+
+## Dataset characteristics
+
+Each dataset is structured with five distinct columns, the most significant being "date" and "value." The "date" column denotes the specific day of data collection.
+In the case of quarterly data, observations are restricted to the first day of January, April, July, and October.
+Notably, FRED does not provide data for August or December quarters.
+This omission is primarily due to practical considerations, as certain data may only become available after the quarter's end, rendering updates before then unproductive.
+
+Furthermore, the "value" column records expenditures in billions of US dollars.
+For a comparative analysis of each category, please refer to @fig-2.
+The cleaned data utilized for modeling and analysis is detailed in @tbl-2.
+The datasheet is accessible in @sec-appen for further reference.
+
+## Measurement
+
+FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S.
+Census Bureau, Bureau of Labor Statistics, and U.S.
+Bureau of Economic Analysis (BEA), among others.
+These datasets are aggregated at the country level.
+We use national-level data because it provides a broader perspective on general economic trends in the U.S.
+All five datasets we have used originate from the BEA.
+The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies.
+Upon receiving these datasets, BEA makes adjustments to the raw survey data.
+For instance, they fill in missing gaps in the datasets.
+For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers.
+Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.
+
+FRED collaborates with both public and private organizations, such as the BEA, to populate its comprehensive database.
+The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S.
+economy, eliminating the need to search various agencies individually for data.
+However, FRED does make further adjustments to BEA datasets before releasing them.
+For example, when aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].
+
+Missing values commonly occur during statutory holidays, when federal offices are closed.
+Consequently, during such weeks, FRED reports data for only six days, excluding the holiday.
+End-of-period calculations are adjusted accordingly, based on the remaining available days within the month, minus one.
+
+All datasets underwent rigorous cleaning and preparation by both FRED and BEA before being made available on the FRED website.
+Consequently, we found no need for extensive cleaning on our part.
+The data was complete with no missing values, and both columns and rows were appropriately labeled and displayed.
+For better readability, we reorganized the columns and consolidated all five datasets into one dataframe, which we used for analysis.
+
+```{r}
+#| label: fig-1
+#| fig-cap: Levels of real consumption expenditure and income expenditure, in billions of dollars, chained to 2017 prices
+#| fig-subcap: ["Food expenditure", "Income"]
+#| echo: false
+#| warning: false
+#| message: false
+#| layout-ncol: 2
+
+data |>
+  ggplot(aes(x = date, y = food_expenditure)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "Date",
+       y = "Changes in real food expenditure",
+       )
+
+data |>
+  ggplot(aes(x = date, y =income_expenditure)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "Date",
+       y = "Changes in real disposable income",
+       )
 ```
@@
        x = "Date",
-       y = "Expenditure",
+       y = "Expenditure (in billions of 2017 USD)",
        fill = "Categories") +
@@
 
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| label: tbl-2
-#| tbl-cap: "Cleaned data showing real expenditure by different categories (in billions of 2017 US dollars)"
-
-columns_to_format <- names(data)[-1]
-
-formatted_data <- data
-formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) sprintf("%.2f", x))
-
-formatted_data[1:12,] |> 
-  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
-                      "Healthcare", "Services"),
-        booktabs = TRUE,
-        linesep = "",
-        align = "c",
-        format.args = list(big.mark = ",")
-        ) |>
-  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
-  kable_styling(font_size = 9, latex_options = "scale_down")  |>
-  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the of the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
-```
-
-```{r}
-#| label: fig-3
-#| fig-cap: Relationship between increases in disposable income and increases in food expenditure, between 2007 and 2022.
-#| echo: false
-#| warning: false
-#| message: false
-
-data |>
-  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
-  geom_point(alpha = 0.5) +
-  geom_smooth(
-    method = "lm",
-    se = TRUE,
-    color = "black",
-    linetype = "dashed",
-    formula = y ~ x
-  ) +
-  theme_classic() + 
-  labs(
-    x = "Real increase in disposable income",
-    y = "Real increase in food expenditure"
-  )
-```
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: tbl-3
-#| tbl-cap: Summary results for both models
-
-first_model <- 
-  readRDS(file = here::here("models/model_1.rds"))
-
-second_model <- 
-  readRDS(file = here::here("models/model_2.rds"))
-
-modelsummary::modelsummary(
-  list(
-    "Model 1" = first_model,
-    "Model 2" = second_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
 ```
@@
 
+```{r}
+#| label: fig-3
+#| fig-cap: Relationship between increases in disposable income and increases in food expenditure, between 2007 and 2022.
+#| echo: false
+#| warning: false
+#| message: false
+
+data |>
+  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(
+    method = "lm",
+    se = TRUE,
+    color = "black",
+    linetype = "dashed",
+    formula = y ~ x
+  ) +
+  theme_classic() + 
+  labs(
+    x = "Real increase in disposable income",
+    y = "Real increase in food expenditure"
+  )
+```
+
 We anticipate a positive correlation between income and food expenditure.
@@
 Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit.
-Notably, the root mean square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.
+Notably, the root-mean-square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.
 
@@
 Finally, healthcare and services expenditure are clustered together.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-3
+#| tbl-cap: Summary results simple linear regression and multiple linear linear regression.
+
+first_model <- 
+  readRDS(file = here::here("models/model_1.rds"))
+
+second_model <- 
+  readRDS(file = here::here("models/model_2.rds"))
+
+modelsummary::modelsummary(
+  list(
+    "Model 1" = first_model,
+    "Model 2" = second_model
+  ),
+  statistic = "mad",
+  fmt = 2
+)
+```
 
@@
 Current research suggests that higher income tends to result in increased expenditure on durable goods, surpassing that of other consumable products.
-Consumers often allocate the majority of their additional income towards items that provide utility over an extended period rather than those consumed in a single instance.
+Consumers often allocate the majority of their additional income towards items that provide utility over an extended period, rather than those consumed in a single instance.
 Using a linear probability function, @lee1964income showed that consumers typically increase their spending on durable goods in response to income growth and resist reducing it even when income decreases.
@@
 This approach aims to enhance the model's external validity and clarify the estimand we are pursuing.
-
-\newpage
 

### 2024-04-16 scripts/02-data_cleaning.R
updated data_cleaning file

@@ -50,3 +50,4 @@ data <- merge(d_f_i, healthcare_services,
 
 #### Save data ####
 write_csv(data, "data/analysis_data/data.csv")
+write_parquet(analysis_data, "data/analysis_data/data.parquet")

### 2024-04-16 paper/paper.qmd
updated the data_simulate file

@@
 The estimand of this paper is the measure of how much food spending changes (in billions of USD) when disposable income increases by a billion USD.
-
-In @sec-data, we discuss the data we used for this project, including information of the software and packages used.
-This is followed any a explanation of the models that we used, which can be found in @sec-model.
-In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income.
-We finish the project with weaknesses and potential areas of improvement in the future.
 
@@
 This validates our hypothesis, suggesting a direct correlation between income and food consumption.
-According to the models, this translates to an economic significance of $70,000,000/$1,000,000,000, equivalent to a 7% increase.
+According to the models, this translates to an economic significance of $20,000,000/$1,000,000,000, equivalent to a 2% increase.
 In the following section, we talk about the reasons behind relatively low economic significance, exploring the impact of savings, consumer behavior and elastiticies, followed by potential weaknesses in our paper.
@@
 
-The income and price elasticities of food are crucial factors in understanding why there's not much change in food spending despite changes in income.
+The income and price elasticities of food are crucial factors in understanding why there is not much change in food spending despite changes in income.
 Income elasticity measures how much food demand shifts when income changes, while price elasticity assesses the impact of price changes on food demand.
@@
 In essence, these findings show that there's only a modest increase in food demand as income rises and a moderate decrease as food prices go up.
-Consequently, changes in income don't lead to significant shifts in food demand.
+This suggests changes in income do nott lead to significant shifts in food demand.
 
@@
 The unique nature of the pandemic's shocks further limits the model's generalizability, especially considering variations in social benefit policies and lockdown measures among different countries.
-Consequently, applying our model to another country within the same timeframe may yield unreliable results due to differing policy responses to COVID-19.
+Consequently, applying our model to another country within the same time frame may yield unreliable results due to differing policy responses to COVID-19.
 
-In future iterations, a narrower timeframe excluding such large-scale global shocks would be prudent.
+In future iterations, a narrower time frame excluding such large-scale global shocks would be prudent.
 This approach aims to enhance the model's external validity and clarify the estimand we are pursuing.

### 2024-04-16 scripts/00-simulate_data.R

@@ -13,18 +13,25 @@ library(tidyverse)
 
 #### Simulate data ####
 
-set.seed(37373)
+set.seed(37373) # For replication purposes
 
-dates <- seq(as.Date("2007-01-01"), by = "quarter", length.out = 60)
 
-# Without controls
+
+# As discussion in the main paper, we look quaterly economic data from 2007 to 2022, which is a 15-year span, 
+# with 60 months totals. FRED releases quaterly data on the first on each quater months, hence, all dates being on the 1st.
+
+dates <- seq(as.Date("2007-01-01"), by = "quarter", length.out = 60) 
+
+# We simulate the primary outcome and predictor variables, food and income. We simulate with data ~ N(950,150). These are 
+# arbitrary values. 
 sim_data <- tibble(
   Date = dates,
   food = rnorm(60, mean = 950, sd = 150),
   income = rnorm(60, mean = 1350, sd = 150)
 )
 
-# Adding in controls
+# We simulate the control variables. these include disposable/disposable, services and healthcare spending.
+# Again, these normal distribution and the values do not signify the actual data. 
 
 sim_data_controls <- tibble(
   Date = dates,

### 2024-04-17 scripts/05-interactive_time_series.R
added interactive plot

@@ -1,0 +1,24 @@
+library(dplyr)
+library(leaflet)
+library(plotly)
+
+
+data <- read_csv(here::here("data/analysis_data/data.csv"))
+
+
+p <- plot_ly(data, x = ~date) %>%
+  add_lines(y = ~durable_expenditure, name = "Durable Expenditure", line = list(color = '#17BECF')) %>%
+  add_lines(y = ~nondurable_expenditure, name = "Nondurable Expenditure", line = list(color = '#7F7F7F')) %>%
+  add_lines(y = ~food_expenditure, name = "Food Expenditure", line = list(color = '#FF7F0E')) %>%
+  add_lines(y = ~income_expenditure, name = "Disposable income", line = list(color = '#a52a2a')) %>%
+  add_lines(y = ~healthcare_expenditure, name = "Healthcare Expenditure", line = list(color = '#cd5c5c')) %>%
+  layout(
+    title = "Time Series Plot (2007-2022)",
+    xaxis = list(title = "Date"),
+    yaxis = list(title = "Expenditure (in billions of 2017 USD)"),
+    hovermode = "x"
+  )
+
+p
+
+htmlwidgets::saveWidget(widget = p, file = "interactive_plot.html")

### 2024-04-17 scripts/05-interactive_time_series.R
deleted interactive map

@@ -21,4 +21,14 @@ p <- plot_ly(data, x = ~date) %>%
 
 p
 
-htmlwidgets::saveWidget(widget = p, file = "interactive_plot.html")
+save_html(p, file = "standalone_plot.html")
+
+ui <- fluidPage(
+  includeHTML("standalone_plot.html")
+)
+
+
+server <- function(input, output, session) {}
+
+
+shinyApp(ui, server)

### 2024-04-17 paper/paper.qmd
added interactive plot
updated the qmd file with interactive map.

@@
 For editing texts, we used `rspell` library [@rspell].
+We created interactive map using `plotly` [@plotly] library and published through `shiny` library [@shiny].
 
@@
   kable_styling(font_size = 9, latex_options = "scale_down")  |>
-  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
+  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the aggregated clean data.
+Including the entire dataset would be redundant and unnecessarily clutter the display.
+To get a better understanding of how the data looks like in entirety, please check out the interactive graph.
+The link can be found on the Github page"))
 ```
@@
 By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods.
-Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames
+Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames.
 
@@
 The datasheet is accessible in @sec-appen for further reference.
+You can find an interactive plot showing the evolution of each expenditure variable on our GitHub page.
 
@@
 
-We use ran trace plots for both models and we are interested in understanding whether the algorithm ran into any issues [@Alexander_2023].
+We ran trace plots for both models as we were interested in understanding whether the algorithm ran into any issues [@Alexander_2023].
 Looking at the plots for both models, we do not encounter any out of the ordinary events.

### 2024-04-17 scripts/05-interactive_time_series.R

@@ -1,6 +1,6 @@
 library(dplyr)
-library(leaflet)
 library(plotly)
+library(shiny)
 
 
 data <- read_csv(here::here("data/analysis_data/data.csv"))
@@ -8,23 +8,23 @@ data <- read_csv(here::here("data/analysis_data/data.csv"))
 
 p <- plot_ly(data, x = ~date) %>%
   add_lines(y = ~durable_expenditure, name = "Durable Expenditure", line = list(color = '#17BECF')) %>%
-  add_lines(y = ~nondurable_expenditure, name = "Nondurable Expenditure", line = list(color = '#7F7F7F')) %>%
+  add_lines(y = ~nondurable_expenditure, name = "Nondurable Expenditure", line = list(color = '#7cfc00')) %>%
   add_lines(y = ~food_expenditure, name = "Food Expenditure", line = list(color = '#FF7F0E')) %>%
   add_lines(y = ~income_expenditure, name = "Disposable income", line = list(color = '#a52a2a')) %>%
-  add_lines(y = ~healthcare_expenditure, name = "Healthcare Expenditure", line = list(color = '#cd5c5c')) %>%
+  add_lines(y = ~healthcare_expenditure, name = "Healthcare Expenditure", line = list(color = '#00008b')) %>%
   layout(
     title = "Time Series Plot (2007-2022)",
     xaxis = list(title = "Date"),
     yaxis = list(title = "Expenditure (in billions of 2017 USD)"),
-    hovermode = "x"
+    hovermode = "x unified"
   )
 
 p
 
-save_html(p, file = "standalone_plot.html")
+save_html(p, file = "other/sketches/interactive_time_series.html")
 
 ui <- fluidPage(
-  includeHTML("standalone_plot.html")
+  includeHTML("interactive_time_series.html")
 )
 
 

### 2024-04-18 paper/paper.qmd
updated the title

@@
 ---
-title: "You have more money in your wallet. This does not mean you will spend extra on food"
+title: "Income's Impact on Food Spending "
 subtitle: "Examining the relationship between disposable income and food expenditure for American consumers from 2007-2022"
@@
 library(fredr)
+library(DiagrammeR)
 
@@
 
-We utilize two linear models in our analysis: simple linear regression and multiple regression.
-Simple linear regression enables us to easily quantify the linear association between food expenditure and disposable income, as illustrated by @fig-3.
+We employ two linear models in our analysis: simple linear regression and multiple regression.
+Simple linear regression provides a straightforward way to quantify the linear relationship between food expenditure and disposable income, as depicted in @fig-3.
 However, this model has limitations.
-Most notably, it does not account for confounding variables, and it requires us to make unrealistic assumptions about the independence of the error term ($\epsilon$) and the predictor variable, income.
-This assumption implies that food expenditure and income are entirely independent of each other, which is problematic.
-Factors like economic downturns, income levels, and other variables can correlate with our model variables.
+Specifically, it does not consider confounding variables and relies on unrealistic assumptions about the independence between the error term ($\epsilon$) and the predictor variable, income.
+This assumption suggests that food expenditure and income are completely independent, which is not the case.
+Variables like economic downturns and income levels can correlate with our model variables.
 
-To address these issues, we employ a multiple linear regression model.
-This model helps us control for potential confounding variables, such as expenditure on non-durable goods, which may influence consumer expenditure on food.
-Using multiple linear regression allows us to better isolate the effect of income on food expenditure.
+To overcome these limitations, we use a multiple linear regression model.
+This approach allows us to control for potential confounding variables, such as expenditure on non-durable goods, which can influence consumer spending on food.
+For instance, when disposable income rises, people might choose to prioritize spending on non-durable goods over food, leading to a decrease in food expenditure.
+This means that non-durable spending directly affects income and indirectly impacts food expenditure—a relationship that simple linear regression fails to capture.
+
+We hypothesize that durable goods, non-durable goods, healthcare spending, and services spending are key confounding variables of interest because they represent a significant portion of consumer spending.
+By employing multiple linear regression, we can account for these variables and more accurately isolate the effect of income on food expenditure.
 

### 2024-04-18 paper/paper.qmd
updated the pdf file

@@
 This has resulted in people having more disposable income, which is the money left after taxes are deducted from one's income.
-Disposable income is crucial to consider as it reflects the actual amount of money available for spending.
+Disposable income is an important factor to consider as it reflects the actual amount of money available for spending.
 Studies indicate that when individuals have higher disposable income, they generally spend more [@friend1946relationship].
@@
 We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007.
-We were interested in a comprehensive 15-year modeling period, which culminated in 2022.
+We are interested in a 15-year modeling period, which culminated in 2022.
 
@@
 For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers.
-Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.
+Therefore, BEA incorporates a separate survey to fill this gap, giving us a wider view of the overall economic conditions.
 
-FRED collaborates with both public and private organizations, such as the BEA, to populate its comprehensive database.
+FRED collaborates with both public and private organizations, such as the BEA, to populate their database.
 The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S.
@@
 
-The income and price elasticities of food are crucial factors in understanding why there is not much change in food spending despite changes in income.
+The income and price elasticities of food are important factors in understanding why there is not much change in food spending despite changes in income.
 Income elasticity measures how much food demand shifts when income changes, while price elasticity assesses the impact of price changes on food demand.

### 2024-04-19 paper/datasheet.qmd
Updated README, Paper(both PDF and QMD) and Datashett

--- 2024-04-06 paper/datasheet_template.qmd
@@
 ---
-title: "Datasheet for data used in the paper, titled 'You have more money in your wallet. Does that mean you will spend extra on food?'"
+title: "Datasheet for data used in the paper, titled 'Income's Impact on Food Spending '"
 author: 

### 2024-04-19 paper/paper.qmd
Updated README, Paper(both PDF and QMD) and Datashett

@@
 In @sec-data, we discuss the data we used for this project, including information of the software and packages used.
-This is followed by a explanation of the models that we used, which can be found in @sec-model.
+This is followed by a explanation on the models that we used, which can be found in @sec-model.
 In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income.
@@
                       among others",
-                      "Durable goods are anything that are generally consumed within a short period of time. Examples
+                      "Non-durable goods are anything that are generally consumed within a short period of time. Examples
                       include food, clothing, cosmetics etc",
-                      "Expenditure in food by all Americans in a time period",
+                      "Expenditure in food by all Americans in a given time period",
                       "Refers to total income that is availble to individuals for consumption after deducing taxes",
@@
 
-We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels.
+We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series data, both at the US national and international levels.
 Utilizing the `fredr` package, we integrated six distinct datasets into one data frame.
-These datasets were:
+These datasets were titled:
 
@@
 All five datasets we have used originate from the BEA.
-The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies.
+The BEA gathers data from over 360 surveys and these collections are sponsored by various Federal and private agencies.
 Upon receiving these datasets, BEA makes adjustments to the raw survey data.
-For instance, they fill in missing gaps in the datasets.
-For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers.
+For example, they fill in missing gaps in the datasets.
+The process requires estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey, Census Bureau’s Monthly Wholesale Trade Survey, does not include inventories from non-merchant sellers.
 Therefore, BEA incorporates a separate survey to fill this gap, giving us a wider view of the overall economic conditions.
@@
 
-In multiple linear regression, we found an estimated intercept of -97.40 billion USD.
-With each $1 billion USD increase in income, food expenditure increased by 20 million USD.
-Similar to the simple regression model, the standard deviation is negligibly small.
+In our multiple linear regression analysis, we estimated an intercept of -$97.40 billion USD.
+For every $1 billion USD rise in income, food expenditure increased by $20 million USD.
+Consistent with the simple regression model, the standard deviation remains insignificantly small.
+
+Durable goods expenditure exhibits a negative relationship with food expenditure, with a decrease of $15 million.
+Similarly, services expenditure sees a decrease of $1 million.
+Thus, the model suggests that an increase in spending on durable goods and services could lead to a decrease in food expenditure.
+Conversely, there is a positive relationship between expenditure on nondurable goods and food, with a notable increase of $47 million in food spending for increases in nondurable goods spending.
+This relationship suggests that these expenditures may be complements of each other.
+Lastly, we observed no significant relationship between healthcare spending and food expenditure.
+Intuitively, this makes sense, as both categories encompass essential items and are less susceptible to changes in the economic environment.
 
 Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit.
-Notably, the root-mean-square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.
+Notably, the root-mean-square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating better predictive accuracy.
 
@@
 A factor that might clarify the low economic significance is that consumers tend to opt for saving their additional income rather than spending it.
-This aligns with existing economic theories, as suggested by [@krugman], which indicate that consumption increases only slightly in response to income growth, as individuals prefer to set aside funds for "rainy day".
+This aligns with existing economic theories, as suggested by @krugman, which indicates that consumption increases only slightly in response to income growth, as individuals prefer to set aside funds for "rainy day".
 This tendency becomes apparent during periods of heightened economic uncertainty, when people tend to prioritize saving as a precautionary measure for the future.
@@
 In essence, these findings show that there's only a modest increase in food demand as income rises and a moderate decrease as food prices go up.
-This suggests changes in income do nott lead to significant shifts in food demand.
+This suggests changes in income do not lead to significant shifts in food demand.
 
@@
 This period was chosen to explore changes in the relationship between food expenditure and income, however, it also captures economic impact of COVID-19, which was not our intention.
-With the pandemic's onset in late 2020, disposable income surged due to social benefit programs (see @fig-1-2), while food expenditure remained relatively stable due to its essential nature (see @fig-1-1).
+With the pandemic's onset in late 2020, disposable income surged due to social benefit programs (see @fig-1-2), while food expenditure remained relatively stable due to it's essential nature (see @fig-1-1).
 This led to erratic swings in income and predictable changes in food spending, distorting our data.
@@
 
-In future iterations, a narrower time frame excluding such large-scale global shocks would be prudent.
-This approach aims to enhance the model's external validity and clarify the estimand we are pursuing.
+In future iterations, a narrower time frame excluding such large-scale global shocks would be more ideal.
+This would enhance the model's external validity and give us an un-biased estimator for our estimand.
 
